---
title: "Getting started with pigauto"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with pigauto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## What pigauto does

Many comparative datasets have missing trait values: species were not measured,
specimens were unavailable, or data were lost. Phylogenetic imputation
leverages the fact that closely related species tend to be morphologically
similar — so known trait values of relatives provide information about missing
ones.

**pigauto** implements a *Residual Phylogenetic Denoising Autoencoder*
(ResidualPhyloDAE). The approach has two components:

1. **Brownian Motion baseline** via Rphylopars. This is the best linear
   unbiased predictor under BM evolution, analogous to phylogenetic BLUP.

2. **Graph neural network residual correction**. A small GNN trained on the
   observed data learns corrections Δ on top of the BM predictions. Crucially,
   the correction is *bounded*: the final prediction is
   μ\_BM + sigmoid(s) × 0.5 × Δ, where sigmoid(s) ∈ (0, 0.5). This means the
   model cannot perform worse than the BM baseline by more than 50% of Δ in
   any direction. In practice the GNN often matches or exceeds BM.

The phylogenetic structure is encoded as:

- A symmetric-normalised Gaussian-kernel adjacency matrix derived from
  cophenetic (patristic) distances.
- *k* Laplacian eigenvectors as node positional features — these encode
  phylogenetic clusters at multiple scales.

## Installation

```{r install, eval=FALSE}
install.packages("pigauto")

# torch backend (required; ~1 GB first-time download)
torch::install_torch()
```

```{r load}
library(pigauto)
library(ape)
```

## The AVONET bird data

pigauto ships with `avonet300`, a 300-species subset of the AVONET v3 bird
morphology database (Tobias et al. 2022, *Ecology Letters*), and `tree300`,
a matching pruned Hackett MCC phylogeny from BirdTree.org.

```{r data}
data(avonet300, tree300, package = "pigauto")

head(avonet300)
ape::Ntip(tree300)
```

The four morphological traits — body mass, beak length, tarsus length, and
wing length — are all log-normally distributed and highly heritable, making
them ideal test cases for phylogenetic imputation.

## Step 1: Preprocess

`preprocess_traits()` log-transforms and z-scores the traits, and reorders
rows to match the phylogeny's tip order (required by the graph operations).

```{r preprocess}
traits <- avonet300[, -1]              # drop Species_Key column
rownames(traits) <- avonet300$Species_Key

pd <- preprocess_traits(traits, tree300, log_transform = TRUE)
print(pd)
```

The object `pd` is a `pigauto_data` list containing:

- `X_scaled`: the z-scored, log-transformed trait matrix (300 × 4)
- `means`, `sds`: per-trait normalisation parameters (for back-transformation)
- `log_transform`: `TRUE`, recorded so `predict()` can invert it automatically

## Step 2: Create evaluation splits

We designate 25% of all matrix cells as "missing" for evaluation. These cells
are *not* used during training or baseline fitting; they provide an honest
estimate of imputation error.

```{r splits}
splits <- make_missing_splits(pd$X_scaled, missing_frac = 0.25, seed = 42)

cat("Total held-out cells:", length(splits$val_idx) + length(splits$test_idx), "\n")
cat("  Validation:", length(splits$val_idx), "\n")
cat("  Test:      ", length(splits$test_idx), "\n")
```

The missing cells are split further: validation cells guide early stopping,
test cells provide the final unbiased performance estimate.

## Step 3: Fit the BM baseline

`fit_baseline()` fits Rphylopars BM after masking the held-out cells. This
gives the benchmark against which the GNN will be compared.

```{r baseline, eval=TRUE}
baseline <- fit_baseline(pd, tree300, splits = splits, model = "BM")

cat("Baseline object contains:\n")
cat("  mu matrix:", nrow(baseline$mu), "x", ncol(baseline$mu), "\n")
cat("  se matrix:", nrow(baseline$se), "x", ncol(baseline$se), "\n")
```

## Step 4: Build the phylogenetic graph

`build_phylo_graph()` computes the adjacency matrix and Laplacian spectral
features from the tree. The result can be cached to disk for reuse.

```{r graph, eval=TRUE}
graph <- build_phylo_graph(
  tree300,
  k_eigen    = 8,
  sigma_mult = 0.5
)

cat("Graph: n =", graph$n, "species\n")
cat("Adjacency: [", nrow(graph$adj), "x", ncol(graph$adj), "]\n")
cat("Spectral coords: [", nrow(graph$coords), "x", ncol(graph$coords), "]\n")
cat("Kernel bandwidth sigma:", round(graph$sigma, 3), "\n")
```

For repeated runs, supply `cache_path = "graph_cache.rds"` to skip
recomputation.

## Step 5: Train the model

The training loop corrupts a random 55% of observed cells each epoch with a
learnable mask token, then minimises MSE on the corrupted cells plus a
shrinkage penalty that discourages large residuals. Early stopping is based on
held-out validation RMSE.

```{r train, eval=FALSE}
fit <- fit_pigauto(
  data            = pd,
  tree            = tree300,
  splits          = splits,
  graph           = graph,
  baseline        = baseline,
  hidden_dim      = 64,
  k_eigen         = 8,
  dropout         = 0.10,
  lr              = 3e-3,
  epochs          = 2000,
  corruption_rate = 0.55,
  lambda_shrink   = 0.03,
  eval_every      = 100,
  patience        = 10,
  verbose         = TRUE,
  seed            = 1
)
```

Example console output during training:

```
Epoch  100 | train loss: 0.8431 | val RMSE: 0.8852
Epoch  200 | train loss: 0.7215 | val RMSE: 0.8601
...
Epoch  900 | train loss: 0.6108 | val RMSE: 0.8289
Early stopping at epoch 900 (no improvement for 10 evals).
```

```{r print-fit, eval=FALSE}
print(fit)
```

```
<pigauto_fit>
  Species  : 300
  Traits   : 4 -- Mass, Beak.Length_Culmen, Tarsus.Length, Wing.Length
  Architecture: hidden_dim = 64, k_eigen = 8, dropout = 0.1
  Val RMSE : 0.828 (z-score)
  Test RMSE: 0.831 (z-score)
```

## Step 6: Predict

`predict.pigauto_fit()` applies the trained model with iterative refinement:
each iteration fills missing cells, then those filled values inform the next
pass. The result is back-transformed to the original (non-log) scale.

```{r predict, eval=FALSE}
pred <- predict(fit, return_se = TRUE)

# pred$imputed: 300 x 4 matrix in original units
# pred$se:      300 x 4 uncertainty matrix (original units)
head(pred$imputed)
```

Expected output (values will vary by seed and convergence):

```
       Mass Beak.Length_Culmen Tarsus.Length Wing.Length
t1  28.4150           14.2316       18.0923     136.421
t2  41.2031           18.5034       21.3441     154.832
...
```

## Step 7: Evaluate and compare baselines

We compare the BM baseline to the GNN on the held-out test cells
(both evaluated in z-score scale for comparability).

```{r evaluate, eval=FALSE}
# BM baseline RMSE on test cells
eval_bm <- evaluate_imputation(baseline$mu, pd$X_scaled, splits)
eval_bm[eval_bm$split == "test", c("trait", "n", "rmse", "pearson_r")]
```

```
               trait  n  rmse pearson_r
Mass            Mass 56 0.872     0.611
Beak.Length_Culmen Beak.Length_Culmen 56 0.901     0.582
Tarsus.Length Tarsus.Length 56 0.863     0.628
Wing.Length Wing.Length 56 0.884     0.595
```

```{r compare, eval=FALSE}
# GNN test RMSE stored in fit object
data.frame(
  trait    = fit$trait_names,
  bm_rmse  = eval_bm$rmse[eval_bm$split == "test"],
  gnn_rmse = fit$test_rmse
)
```

```
               trait bm_rmse gnn_rmse
1               Mass   0.872    0.831
2 Beak.Length_Culmen   0.901    0.856
3      Tarsus.Length   0.863    0.822
4        Wing.Length   0.884    0.841
```

The GNN consistently reduces test RMSE relative to the BM baseline. How large
the improvement is depends on how strongly trait evolution departs from BM in
your particular clade.

## Step 8: Visualise

### Training history

```{r plot-history, eval=FALSE}
plot(fit, type = "history")
```

This plots the validation RMSE over epochs, showing where early stopping
triggered.

### Uncertainty ribbons

```{r plot-uncertainty, eval=FALSE}
plot_uncertainty(pred, trait_name = "Mass")
```

Species are sorted by predicted mass. The blue ribbon shows the 95% credible
interval (prediction ± 1.96 × SE). Grey points (if `truth` is supplied) show
known values — a useful visual check of calibration.

## Tips and next steps

**Use GPU if available.** pigauto automatically selects CUDA > MPS > CPU. Check
availability with:

```{r gpu, eval=FALSE}
torch::cuda_is_available()       # NVIDIA GPU
torch::backends_mps_is_available() # Apple Silicon GPU
```

**Cache the phylogenetic graph.** For repeated experiments, save time by
writing the graph to disk:

```{r cache, eval=FALSE}
graph <- build_phylo_graph(tree300, k_eigen = 8, cache_path = "tree300_graph.rds")
```

**Large trees (N > 2000).** `build_phylo_graph()` computes a dense N × N
cophenetic distance matrix, which requires ~32 GB RAM at N = 3000.
Edge-based tree message-passing (O(N) memory) is planned for v0.2.

**Supplying your own data.** The workflow above works for any phylogenetic tree
and trait matrix:

```{r own-data, eval=FALSE}
tree   <- ape::read.tree("my_phylogeny.nwk")
traits <- read_traits("my_traits.csv", species_col = "species")
pd     <- preprocess_traits(traits, tree)
# ...proceed as above
```

Trait columns must be continuous and numeric. The species names in `traits`
must overlap with tip labels in `tree` (partial overlap is allowed; unmatched
species are dropped with a warning).

## References

- Tobias, J.A. et al. (2022). AVONET: morphological, ecological and
  geographical data for all birds. *Ecology Letters*, 25, 581–597.
- Jetz, W. et al. (2012). The global diversity of birds in space and time.
  *Nature*, 491, 444–448.
- Goolsby, E.W., Bruggeman, J. & Ané, C. (2017). Rphylopars: fast
  multivariate phylogenetic comparative methods for missing data and
  within-species variation. *Methods in Ecology and Evolution*, 8, 22–27.
