---
title: "Getting started with pigauto"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with pigauto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Start here

The beginner workflow reads files, checks them before fitting, and keeps the
completed-data, diagnostic-prediction, uncertainty, and inference roles
separate.

```{r canonical-journey, eval=FALSE}
library(pigauto)

traits <- read_traits("traits.csv")
tree <- read_tree("tree.nwk")
check_pigauto(traits, tree)
result <- impute(traits, tree)
completed <- completed_data(result)
pigauto_report(result)
```

`check_pigauto()` is fit-free. Resolve its structured errors before fitting;
inspect `check$species$matched`, `$data_only`, and `$tree_only` when labels do
not reconcile. `completed` preserves observed values and contains fills only
for modeled missing cells. `result$prediction` is an all-cell diagnostic
output; its uncertainty is type-dependent, and conformal bounds are nominal
held-out diagnostics. A `pigauto_result` does not authorize downstream
inference: use `multi_impute_analysis()` only in its documented narrow regime.

`read_traits()` requires unique species identifiers. For repeated observations,
use `read.csv()` and supply `species_col = "species"` to both the check and
`impute()`.

### Common declarations

```{r declarations, eval=FALSE}
# Ordered ecological states
traits$threat <- ordered(traits$threat, levels = c("LC", "NT", "VU", "EN", "CR"))

# Integer-valued continuous measurement; integer otherwise means count
result <- impute(traits, tree, trait_types = c(length_mm = "continuous"))

# Numeric 0-1 proportion and zero-inflated count require explicit declarations
result <- impute(traits, tree, trait_types = c(cover = "proportion", parasites = "zi_count"))

# A composition is complete by row and its observed components sum to one
result <- impute(traits, tree,
  multi_proportion_groups = list(diet = c("diet_insect", "diet_fruit", "diet_seed")))
```

## What pigauto does

Many comparative datasets have missing trait values: species were not
measured, specimens were unavailable, or data were lost. pigauto fills
in those gaps using three sources of information:

1. **The phylogenetic tree.** Closely related species tend to share
   similar traits, so known values of relatives are informative about
   missing ones.
2. **Cross-trait correlations.** If body mass predicts beak length,
   then observed masses help impute missing beak lengths even when the
   species itself was never measured for that trait.
3. **Environmental covariates** (optional). Climate, habitat, or
   geographic variables can improve imputation when trait variation
   has a strong environmental component beyond phylogenetic signal.

The package handles continuous measurements, counts, binary variables,
ordered categories, unordered categories, bounded proportions,
zero-inflated counts, and compositional multi-proportion data — all in a
single call to `impute()`.

### How it works (briefly)

pigauto combines a phylogenetic baseline with a graph neural network
(GNN) correction. The baseline uses Brownian-motion conditional
imputation for continuous-family latent columns and phylogenetic
label-propagation or threshold/liability candidates for discrete-family
columns. The GNN learns
additional patterns from tree topology and cross-trait structure. A
per-trait gate controls how much the GNN contributes: when the baseline
is already good enough, the gate closes and the GNN stays out of the
way. This is verified on held-out data after training.

> **Technical note.** The internal class name `ResidualPhyloDAE` refers
> to the ResNet-style skip connections inside the GNN layers, not to a
> statistical residual `y - baseline`.

The phylogenetic structure is encoded as:

- A symmetric-normalised Gaussian-kernel adjacency matrix derived from
  cophenetic (patristic) distances.
- *k* Laplacian eigenvectors as node positional features — these encode
  phylogenetic clusters at multiple scales.

## Installation

```{r install, eval=FALSE}
# CRAN release
install.packages("pigauto")

# Development version
pak::pak("itchyshin/pigauto")

# torch backend (required; ~1 GB first-time download)
torch::install_torch()
```

```{r load}
library(pigauto)
library(ape)
```

## Inspecting the internal pipeline

The six-line journey above is the primary interface. The following is the
fine-grained pipeline for inspection, caching, or benchmarking.

```{r quickstart, eval=FALSE}
result <- impute(traits, tree)

result$completed    # data.frame: observed values preserved, NAs filled
result$imputed_mask # logical matrix: TRUE where a cell was imputed
result$prediction$se # per-cell uncertainty (original units)
```

`impute()` runs the full pipeline internally — preprocessing, baseline fitting,
graph construction, GNN training, calibration, and prediction. The rest of this
vignette walks through each step individually for users who need fine-grained
control, want to cache intermediate objects, or are benchmarking components.

## The AVONET bird data

pigauto ships with `avonet300`, a 300-species subset of the AVONET v3 bird
morphology database (Tobias et al. 2022, *Ecology Letters*), and `tree300`,
a matching pruned Hackett MCC phylogeny from BirdTree.org.

```{r data}
data(avonet300, tree300, package = "pigauto")

head(avonet300)
ape::Ntip(tree300)
```

The four morphological traits — body mass, beak length, tarsus length, and
wing length — are all log-normally distributed and highly heritable, making
them ideal test cases for phylogenetic imputation.

## Step 1: Preprocess

`preprocess_traits()` log-transforms and z-scores the traits, and reorders
rows to match the phylogeny's tip order (required by the graph operations).

```{r preprocess}
traits <- avonet300[, -1]              # drop Species_Key column
rownames(traits) <- avonet300$Species_Key

pd <- preprocess_traits(traits, tree300, log_transform = TRUE)
print(pd)
```

The object `pd` is a `pigauto_data` list containing:

- `X_scaled`: the z-scored, log-transformed trait matrix (300 × 4)
- `means`, `sds`: per-trait normalisation parameters (for back-transformation)
- `log_transform`: `TRUE`, recorded so `predict()` can invert it automatically

## Step 2: Create evaluation splits

We designate 25% of all matrix cells as "missing" for evaluation. These cells
are *not* used during training or baseline fitting; they provide an honest
estimate of imputation error.

```{r splits}
splits <- make_missing_splits(pd$X_scaled, missing_frac = 0.25, seed = 42)

cat("Total held-out cells:", length(splits$val_idx) + length(splits$test_idx), "\n")
cat("  Validation:", length(splits$val_idx), "\n")
cat("  Test:      ", length(splits$test_idx), "\n")
```

The missing cells are split further: validation cells guide early stopping,
test cells provide the final unbiased performance estimate.

## Step 3: Fit the BM baseline

`fit_baseline()` fits the phylogenetic BM baseline after masking the held-out cells. This
gives the benchmark against which the GNN will be compared.

```{r baseline, eval=TRUE}
baseline <- fit_baseline(pd, tree300, splits = splits, model = "BM")

cat("Baseline object contains:\n")
cat("  mu matrix:", nrow(baseline$mu), "x", ncol(baseline$mu), "\n")
cat("  se matrix:", nrow(baseline$se), "x", ncol(baseline$se), "\n")
```

## Step 4: Build the phylogenetic graph

`build_phylo_graph()` computes the adjacency matrix and Laplacian spectral
features from the tree. The result can be cached to disk for reuse.

```{r graph, eval=TRUE}
graph <- build_phylo_graph(
  tree300,
  k_eigen    = 8,
  sigma_mult = 0.5
)

cat("Graph: n =", graph$n, "species\n")
cat("Adjacency: [", nrow(graph$adj), "x", ncol(graph$adj), "]\n")
cat("Spectral coords: [", nrow(graph$coords), "x", ncol(graph$coords), "]\n")
cat("Kernel bandwidth sigma:", round(graph$sigma, 3), "\n")
```

For repeated runs, supply `cache_path = "graph_cache.rds"` to skip
recomputation.

## Step 5: Train the model

The training loop corrupts a random 55% of observed cells each epoch with a
learnable mask token, then minimises the type-appropriate loss on the
corrupted cells plus a shrinkage penalty on `delta - baseline` that
pulls the GNN toward the baseline. Early stopping is based on held-out
validation RMSE.

```{r train, eval=FALSE}
fit <- fit_pigauto(
  data            = pd,
  tree            = tree300,
  splits          = splits,
  graph           = graph,
  baseline        = baseline,
  hidden_dim      = 64,
  k_eigen         = 8,
  dropout         = 0.10,
  lr              = 3e-3,
  epochs          = 2000,
  corruption_rate = 0.55,
  lambda_shrink   = 0.03,
  eval_every      = 100,
  patience        = 10,
  verbose         = TRUE,
  seed            = 1
)
```

Example console output during training:

```
Epoch  100 | train loss: 0.8431 | val RMSE: 0.8852
Epoch  200 | train loss: 0.7215 | val RMSE: 0.8601
...
Epoch  900 | train loss: 0.6108 | val RMSE: 0.8289
Early stopping at epoch 900 (no improvement for 10 evals).
```

```{r print-fit, eval=FALSE}
print(fit)
```

```
<pigauto_fit>
  Species  : 300
  Traits   : 4 -- Mass, Beak.Length_Culmen, Tarsus.Length, Wing.Length
  Architecture: hidden_dim = 64, k_eigen = 8, dropout = 0.1
  Val RMSE : 0.828 (z-score)
  Test RMSE: 0.831 (z-score)
```

## Step 6: Predict

`predict.pigauto_fit()` runs a single forward pass through the trained model,
optionally repeated with MC dropout to generate multiple stochastic completions
for uncertainty quantification. The result is back-transformed to the original
(non-log) scale.

```{r predict, eval=FALSE}
pred <- predict(fit, return_se = TRUE)

# pred$imputed: 300 x 4 matrix in original units
# pred$se:      300 x 4 uncertainty matrix (original units)
head(pred$imputed)

# Conformal prediction intervals (nominal held-out diagnostic)
pred$conformal_lower[["Mass"]]    # lower bound per species (original units)
pred$conformal_upper[["Mass"]]    # upper bound per species (original units)
pred$conformal_coverage           # empirical coverage on val set (target ≈ 0.95)
```

Expected output (values will vary by seed and convergence):

```
       Mass Beak.Length_Culmen Tarsus.Length Wing.Length
t1  28.4150           14.2316       18.0923     136.421
t2  41.2031           18.5034       21.3441     154.832
...
```

## Step 7: Evaluate — and read the gate

`evaluate_imputation()` scores held-out cells on the scale that matches
the trait type. On **high-signal morphometrics** (AVONET-like mass,
beak, tarsus, wing) the calibrated gate often **closes**
(`r_cal = 0`). Then the shipped prediction *is* the phylogenetic
baseline; the GNN is a no-op, not a guaranteed improvement.

```{r evaluate, eval=FALSE}
# Held-out scores (z-score RMSE for continuous traits)
evaluate_imputation(pred$imputed_latent, pd$X_scaled, splits)

# Per-trait calibrated gate. 0 means baseline-only.
fit$r_cal
```

Do not treat a GNN-versus-BM table as the default story. If the gate is
closed, a lower GNN training loss on masked cells is not a claim that
the ensemble beat BM on the test cells you ship. How much (if anything)
the GNN adds depends on the trait, clade, missingness pattern, and
whether calibration left the gate open.

## Step 8: Visualise

### Training history

```{r plot-history, eval=FALSE}
plot(fit, type = "history")
```

This plots the validation RMSE over epochs, showing where early stopping
triggered.

### Uncertainty ribbons

```{r plot-uncertainty, eval=FALSE}
plot(pred, type = "intervals", trait = "Mass")
```

Species are sorted by predicted mass. The blue ribbon shows the conformal
prediction interval when conformal scores are available. Grey points (if
`truth` is supplied) show known values — a useful visual check of calibration.

## Experimental analysis-aware multiple imputation

Downstream inference requires imputations compatible with the substantive
analysis. `multi_impute_analysis()` therefore receives the formula and model
class before drawing one incomplete continuous covariate under MAR.

This validates only the documented narrow scope; the interface remains
experimental.

```{r mi-workflow, eval=FALSE}
analysis_data$z_sq <- analysis_data$z^2

mi <- multi_impute_analysis(
  data = analysis_data,
  formula = y ~ x + z,
  missing = "x",
  model = "lm",
  m = 50L,
  auxiliary = "z_sq",
  seed = 1L
)

fits <- with_imputations(mi, function(d) lm(y ~ x + z, data = d))
pool_mi(fits)
```

The initial dispatch is deliberately narrow: proper Bayesian normal-regression
MI for Gaussian `lm`, `smcfcs` for binomial-logit `glm`, and `jomo.smc` for a
Gaussian `lmer` with one random intercept. Users must precompute auxiliary
terms such as `z_sq`. Fixed effects are the only supported pooling targets.

Conformal-width, Brownian/MC-dropout, and PMM draws remain experimental
prediction diagnostics. They failed the downstream fixed-effect gate and must
not be passed to `with_imputations()`/`pool_mi()` for inference.

## Phylogenetic tree uncertainty

Tree uncertainty was outside the analysis-aware validation campaign.
`multi_impute_trees()` may be used descriptively to examine point-prediction
sensitivity, but its stochastic datasets are unsupported for downstream
inference and cannot currently be combined with `multi_impute_analysis()`.

## Active imputation: where to measure next

`suggest_next_observation()` returns model-based BM/label-propagation proxy
rankings under its stated assumptions. It is not an optimal sampling guarantee
or a demonstrated field gain.

```{r active-impute, eval=FALSE}
res <- impute(traits, tree)

# Top-10 individual (species, trait) cells, ordered by expected
# total variance reduction:
suggest_next_observation(res, top_n = 10, by = "cell")

# Or aggregated by species (sum of variance reductions across the
# species' currently-missing continuous-family traits):
suggest_next_observation(res, top_n = 10, by = "species")
```

The `delta_var_total` column is the model-based expected reduction
in total predictive variance across all currently-missing cells if
you observed that candidate next:

\[
\Delta V(s_{\text{new}}) = \sigma_t^2 \sum_{i \in \mathrm{miss}_t}
\frac{D_{ik}^2}{\alpha_k}
\]

where \(D = R_{mm} - R_{mo} R_{oo}^{-1} R_{om}\) is the residual
matrix at currently-missing cells, \(\alpha_k = D_{kk}\) is the
current relative leverage of cell \(k\), and \(\sigma_t^2\) is
the REML BM variance. Treat it as a proxy ranking conditional on the fitted
model, rather than a claim about a future sampling campaign.

**Scope.** `suggest_next_observation()` v2 (2026-05-01) supports
all eight pigauto trait types:

* **Variance reduction** for continuous-family traits (`continuous`,
  `count`, `ordinal`, `proportion`) and the magnitude column of
  `zi_count`, computed by closed-form Sherman-Morrison rank-1
  update on the BM phylogenetic-correlation matrix.
* **Entropy reduction** for `binary`, `categorical`, and the gate
  column of `zi_count`, computed by closed-form label-propagation
  posteriors.
* **Multi-component variance reduction** (per-component BM,
  summed across the K CLR-z latent columns) for `multi_proportion`.

For `zi_count` rows the output populates BOTH `delta_var_total`
(probability-weighted magnitude variance reduction) AND
`delta_entropy_total` (gate entropy reduction); the `metric`
column is set to `"variance"` so rows sort on the magnitude scale.
See `?suggest_next_observation` for full details.

**Variance vs entropy are not directly comparable** — the cross-
metric ordering by `delta_var_total` is approximate.  When you
want a strict ranking on a single metric, filter by `metric`
first or restrict `types` to continuous-family-only or
discrete-only.

Multi-obs (multiple observations per species) inputs are not
supported and error with a clear message.

## Advanced controls

### Solver, calibration, EM, and refinement

```{r advanced-controls, eval=FALSE}
result <- impute(
  traits, tree,
  joint_solver = "rphylopars",
  em_iterations = 2L,
  joint_refine_iter = 1L,
  conformal_split_val = TRUE
)
```

### Exact conditional prediction route

```{r exact-route, eval=FALSE}
result <- impute(traits, tree, predict_method = "exact")
```

### Phylogenetic-signal gating

`impute()` runs a per-trait Pagel's-λ check before training and
routes traits with very weak phylogenetic signal to the
**grand-mean corner** of the safety-floor simplex (i.e., it
predicts the column mean for those traits).  This protects against
over-fitting noise on traits where the phylogeny carries little
information.

It is on by default since v0.9.1.9003.  The two arguments you can
tune:

```{r phylo-signal-gate, eval=FALSE}
result <- impute(traits, tree,
                 phylo_signal_gate     = TRUE,   # default
                 phylo_signal_threshold = 0.2,   # default; min lambda to keep BM/GNN
                 phylo_signal_method   = "lambda")
```

If a trait's λ falls below `phylo_signal_threshold`, the calibrated
gate routes it to the grand-mean corner.  `print(fit)` reports
which traits triggered the gate and their λ values.

To disable (e.g., when comparing against an older snapshot or for
benchmarking purposes):

```{r phylo-signal-gate-off, eval=FALSE}
result <- impute(traits, tree, phylo_signal_gate = FALSE)
```

### Assembling covariates from public data

Two helpers are bundled for the common case of building a
covariate matrix from species-level environmental data:

```{r covariate-helpers, eval=FALSE}
# Step 1: GBIF occurrence centroids (median lat/lon across cleaned points)
gbif <- pull_gbif_centroids(species          = rownames(my_traits),
                            cache_dir        = "cache/gbif",
                            occurrence_limit = 500L)
#> data.frame with columns species, centroid_lat, centroid_lon, n_occurrences

# Step 2: WorldClim bioclim summaries (median + IQR of all 19 bio variables
# across each species' GBIF occurrences)
clim <- pull_worldclim_per_species(species             = rownames(my_traits),
                                    gbif_cache_dir      = "cache/gbif",
                                    worldclim_cache_dir = "cache/worldclim",
                                    resolution          = "10m")
#> data.frame with bio1..bio19 medians + IQRs + n_extracted per species

# Step 3: hand the whole climate frame to impute() as covariates
result <- impute(my_traits, my_tree, covariates = clim)
```

Caching is on by default — both helpers write to the
`cache_dir` you supply and re-use existing extracts on subsequent
runs.  GBIF requires no API key.  WorldClim raster downloads are
~50–500 MB depending on resolution and are cached in
`worldclim_cache_dir`.  Both functions error with a clear message
if the optional dependencies (`rgbif`, `terra`) are not installed.
Both helpers return one row per requested species; alignment is
enforced by name where names are present, so pass `species =` and
keep the returned `species` column intact rather than reordering it.

For a worked end-to-end example with covariates, use this section alongside
`?impute`, `?pull_gbif_centroids`, and `?pull_worldclim_per_species`. The old
static covariate walk-through is being refreshed and is not the current source
of truth.

## Tips and next steps

**Use GPU if available.** pigauto automatically selects CUDA > MPS > CPU. Check
availability with:

```{r gpu, eval=FALSE}
torch::cuda_is_available()       # NVIDIA GPU
torch::backends_mps_is_available() # Apple Silicon GPU
```

**Cache the phylogenetic graph.** For repeated experiments, save time by
writing the graph to disk:

```{r cache, eval=FALSE}
graph <- build_phylo_graph(tree300, k_eigen = 8, cache_path = "tree300_graph.rds")
```

**Large trees (N > 2000).** `build_phylo_graph()` computes a dense N × N
cophenetic distance matrix and a Laplacian eigendecomposition. The eigendecomposition
is O(N³) in the worst case, but `k_eigen = "auto"` caps the number of eigenvectors
at 32 (the default for all trees), which keeps computation tractable in practice —
pigauto has been tested up to 10,000 species on a standard laptop. The main
cost at very large N is memory for the N × N distance matrix (~800 MB at
N = 10,000). If memory is tight, prune to a clade of interest or watch for a
future release that switches to a sparse Lanczos eigensolver.

**Supplying your own data.** The workflow above works for any phylogenetic tree
and trait matrix:

```{r own-data, eval=FALSE}
tree   <- ape::read.tree("my_phylogeny.nwk")
traits <- read_traits("my_traits.csv", species_col = "species")
pd     <- preprocess_traits(traits, tree)
# ...proceed as above
```

Trait columns can be any of eight supported types. Five are
auto-detected from R column class: `numeric` (continuous), `integer`
(count), `factor` with 2 levels (binary), `factor` with >2 levels
(categorical), and `ordered` (ordinal). Three require an explicit
declaration because they cannot be inferred from class alone:
`numeric` (0–1) with `trait_types = "proportion"` for bounded
proportions; `integer` with `trait_types = "zi_count"` for
zero-inflated counts; and K `numeric` columns declared via the
`multi_proportion_groups` argument for compositional (simplex) data
whose rows sum to 1. pigauto reads column classes to decide how each
trait is encoded and imputed. The
species names in `traits` must overlap with tip labels in `tree` (partial
overlap is allowed; unmatched species are dropped with a warning).

## References

- Tobias, J.A. et al. (2022). AVONET: morphological, ecological and
  geographical data for all birds. *Ecology Letters*, 25, 581–597.
- Jetz, W. et al. (2012). The global diversity of birds in space and time.
  *Nature*, 491, 444–448.
- Goolsby, E.W., Bruggeman, J. & Ané, C. (2017). Rphylopars: fast
  multivariate phylogenetic comparative methods for missing data and
  within-species variation. *Methods in Ecology and Evolution*, 8, 22–27.
  (pigauto's internal BM baseline implements the same conditional-MVN
  approach.)
- Cohn, D.A., Ghahramani, Z. & Jordan, M.I. (1996). Active learning
  with statistical models. *Journal of Artificial Intelligence Research*,
  4, 129–145. (Variance-reduction framework that
  `suggest_next_observation()` adapts to the phylogenetic setting.)
