---
title: "Mixed-Type Trait Imputation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed-Type Trait Imputation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = torch::torch_is_installed()
)
```

## Overview

Real comparative datasets contain many kinds of traits: body mass
(continuous), clutch size (count), migratory status (yes/no), diet type
(carnivore/herbivore/omnivore), threat status (LC < VU < EN). pigauto
handles all of them in a single model — you do not need separate
imputation runs for different column types.

| Type | R class | Example | Notes |
|------|---------|---------|-------|
| Continuous | `numeric` | Body mass, wing length | Auto-detected |
| Count | `integer` | Clutch size, litter size | Auto-detected |
| Binary | `factor` (2 levels) | Migratory yes/no | Auto-detected |
| Categorical | `factor` (>2 levels) | Diet type, lifestyle | Auto-detected |
| Ordinal | `ordered` | Threat status (LC < VU < EN) | Auto-detected |
| Proportion | `numeric` (0–1) | Habitat cover, diet fraction | Requires `trait_types = "proportion"` override |
| ZI count | `integer` (zero-inflated) | Parasite load, rare behaviour | Requires `trait_types = "zi_count"` override; experimental — accuracy more variable than other types |

Each type is automatically detected from the R column class and encoded
into a shared latent space. The phylogenetic baseline and GNN correction
both operate in this space; type-specific logic appears only at encoding,
loss computation, and decoding.

## Synthetic example

```{r simulate}
library(pigauto)
library(ape)

set.seed(42)
n <- 60
tree <- rtree(n)

traits <- data.frame(
  row.names = tree$tip.label,
  mass      = exp(rnorm(n, 3, 0.5)),
  clutch    = as.integer(rpois(n, 3) + 1L),
  migr      = factor(sample(c("no", "yes"), n, replace = TRUE)),
  diet      = factor(sample(c("herb", "carn", "omni"), n, replace = TRUE)),
  threat    = ordered(sample(c("LC", "VU", "EN"), n, replace = TRUE),
                      levels = c("LC", "VU", "EN"))
)
```

## Preprocessing

`preprocess_traits()` auto-detects column types from R classes:

```{r preprocess}
pd <- preprocess_traits(traits, tree)
print(pd)
```

The `trait_map` records each trait's type, levels, latent column
indices, and normalisation parameters:

```{r trait_map}
str(pd$trait_map, max.level = 1)
pd$trait_map$diet
```

## Creating splits

When `trait_map` is supplied, `make_missing_splits()` operates at the
original-trait level.  For categorical traits, all K one-hot columns
are held out together:

```{r splits}
spl <- make_missing_splits(pd$X_scaled, missing_frac = 0.20,
                           seed = 1, trait_map = pd$trait_map)
cat("Val cells (latent):", length(spl$val_idx), "\n")
cat("Test cells (latent):", length(spl$test_idx), "\n")
```

## Baseline fitting

The BM baseline uses phylogenetic conditional MVN for continuous, count,
and ordinal traits.  Binary traits receive the logit of the observed proportion;
categorical traits receive log marginal frequencies:

```{r baseline}
bl <- fit_baseline(pd, tree, splits = spl)
dim(bl$mu)
```

## Training

`fit_pigauto()` uses type-specific losses and trait-level corruption
masking automatically when a `trait_map` is present:

```{r train, message = FALSE}
fit <- fit_pigauto(
  pd, tree,
  splits = spl,
  epochs = 200L,
  eval_every = 50L,
  patience = 5L,
  verbose = FALSE,
  seed = 1
)
print(fit)
```

## Prediction and decoding

`predict()` decodes latent predictions back to original types:

```{r predict}
pred <- predict(fit, return_se = TRUE)
head(pred$imputed)
```

For binary and categorical traits, class probabilities are available:

```{r probs}
# Binary: probability of "yes"
head(pred$probabilities$migr)

# Categorical: probability of each diet class
head(pred$probabilities$diet)
```

The SE matrix provides type-appropriate uncertainty:

```{r se}
head(pred$se)
```

## Evaluation

`evaluate_imputation()` dispatches type-specific metrics:

```{r evaluate}
ev <- evaluate_imputation(pred, pd$X_scaled, spl)
print(ev)
```

## Multiple imputation for downstream inference

The recommended workflow uses `multi_impute()` to generate M complete datasets
from the model's calibrated uncertainty distribution, then pools downstream
coefficients with Rubin's rules via `with_imputations()` and `pool_mi()`.

In **v0.6.1**, `draws_method = "conformal"` is the default: missing cells are
drawn from Normal distributions centred on the point estimate with width set by
the split-conformal calibration score — properly calibrated against held-out
residuals. The alternative `draws_method = "mc_dropout"` runs M stochastic
GNN forward passes with dropout active and BM posterior draws as the blend
baseline; it is available for comparison.

```{r mi_workflow, eval=FALSE}
# Generate M = 20 stochastic complete datasets
mi <- multi_impute(fit$data$X_original, tree, m = 20L)

# Fit a downstream model to each
fits <- with_imputations(mi, function(d) {
  glm(Migratory ~ log(Mass) + Diet, data = d, family = binomial)
})

# Pool with Rubin's rules
pool_mi(fits)
```

`pool_mi()` returns a tidy data.frame with `estimate`, `std.error`, `p.value`,
`df` (Barnard-Rubin degrees of freedom), `fmi` (fraction of missing information),
and `riv` (relative increase in variance) per coefficient.

For reference, the lower-level `predict(fit, n_imputations = 5L)` interface
returns M complete datasets directly (without downstream pooling):
