---
title: "Mixed-Type Trait Imputation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed-Type Trait Imputation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = torch::torch_is_installed()
)
```

## Overview

Real comparative datasets contain many kinds of traits: body mass
(continuous), clutch size (count), migratory status (yes/no), diet type
(carnivore/herbivore/omnivore), threat status (LC < VU < EN). pigauto
handles all of them in a single model — you do not need separate
imputation runs for different column types.

| Type | R class | Example |
|------|---------|---------|
| Continuous | `numeric` | Body mass, wing length |
| Count | `integer` | Clutch size, litter size |
| Binary | `factor` (2 levels) | Migratory yes/no |
| Categorical | `factor` (>2 levels) | Diet type, lifestyle |
| Ordinal | `ordered` | Threat status (LC < VU < EN) |

Each type is automatically detected from the R column class and encoded
into a shared latent space. The phylogenetic baseline and GNN correction
both operate in this space; type-specific logic appears only at encoding,
loss computation, and decoding.

## Synthetic example

```{r simulate}
library(pigauto)
library(ape)

set.seed(42)
n <- 60
tree <- rtree(n)

traits <- data.frame(
  row.names = tree$tip.label,
  mass      = exp(rnorm(n, 3, 0.5)),
  clutch    = as.integer(rpois(n, 3) + 1L),
  migr      = factor(sample(c("no", "yes"), n, replace = TRUE)),
  diet      = factor(sample(c("herb", "carn", "omni"), n, replace = TRUE)),
  threat    = ordered(sample(c("LC", "VU", "EN"), n, replace = TRUE),
                      levels = c("LC", "VU", "EN"))
)
```

## Preprocessing

`preprocess_traits()` auto-detects column types from R classes:

```{r preprocess}
pd <- preprocess_traits(traits, tree)
print(pd)
```

The `trait_map` records each trait's type, levels, latent column
indices, and normalisation parameters:

```{r trait_map}
str(pd$trait_map, max.level = 1)
pd$trait_map$diet
```

## Creating splits

When `trait_map` is supplied, `make_missing_splits()` operates at the
original-trait level.  For categorical traits, all K one-hot columns
are held out together:

```{r splits}
spl <- make_missing_splits(pd$X_scaled, missing_frac = 0.20,
                           seed = 1, trait_map = pd$trait_map)
cat("Val cells (latent):", length(spl$val_idx), "\n")
cat("Test cells (latent):", length(spl$test_idx), "\n")
```

## Baseline fitting

The BM baseline uses Rphylopars for continuous, count, and ordinal
traits.  Binary traits receive the logit of the observed proportion;
categorical traits receive log marginal frequencies:

```{r baseline}
bl <- fit_baseline(pd, tree, splits = spl)
dim(bl$mu)
```

## Training

`fit_pigauto()` uses type-specific losses and trait-level corruption
masking automatically when a `trait_map` is present:

```{r train, message = FALSE}
fit <- fit_pigauto(
  pd, tree,
  splits = spl,
  epochs = 200L,
  eval_every = 50L,
  patience = 5L,
  verbose = FALSE,
  seed = 1
)
print(fit)
```

## Prediction and decoding

`predict()` decodes latent predictions back to original types:

```{r predict}
pred <- predict(fit, return_se = TRUE)
head(pred$imputed)
```

For binary and categorical traits, class probabilities are available:

```{r probs}
# Binary: probability of "yes"
head(pred$probabilities$migr)

# Categorical: probability of each diet class
head(pred$probabilities$diet)
```

The SE matrix provides type-appropriate uncertainty:

```{r se}
head(pred$se)
```

## Evaluation

`evaluate_imputation()` dispatches type-specific metrics:

```{r evaluate}
ev <- evaluate_imputation(pred, pd$X_scaled, spl)
print(ev)
```

## Multiple imputation (MC dropout)

Set `n_imputations > 1` to run M stochastic forward passes with
dropout active.  This produces M complete datasets for proper multiple
imputation:

```{r mc_dropout}
pred_mc <- predict(fit, n_imputations = 5L)
cat("Imputed datasets:", length(pred_mc$imputed_datasets), "\n")
head(pred_mc$se)  # between-imputation SD
```
