---
title: "Multiple imputation and downstream models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple imputation and downstream models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This article shows how to carry imputation uncertainty into a downstream
model: impute the missing cells `m` times, fit the same model to each
completed dataset, and combine the fits. The chunks are not evaluated when
the article is built because the posterior sampler takes minutes. They were
run end to end once with shortened sampler chains.

## Which imputations can be pooled

`multi_impute()` offers three ways to draw the missing cells. Only one of
them is a proper multiple imputation for downstream inference.

| `draws_method` | Use it for | Pass to `with_imputations()` and `pool_mi()`? |
|---|---|---|
| `"posterior"` | Downstream inference on continuous traits | Yes |
| `"conformal"` (current default) | Looking at the spread of plausible values | No, refused |
| `"mc_dropout"` | Looking at the spread of plausible values | No, refused |

The conformal and MC-dropout draws perturb each missing cell around a point
prediction. They do not draw the cells jointly from their distribution given
the observed data. In a simulation of a phylogenetic regression slope (16
regimes: two traits, n = 300 or 1000, Pagel's lambda 1 or 0.5, random or
clade-structured missingness, 120 replicates each), conformal draws biased
the pooled slope towards zero by 0.20 to 0.46 and the pooled 95% intervals
covered the true slope in 0 to 17% of replicates. Other estimands were not
measured. `with_imputations()` therefore refuses these draws. A list of fits
built by hand is still pooled by `pool_mi()`, with a warning that its
provenance cannot be checked; do not do this with conformal or MC-dropout
draws.

This applies to the draws, not to the per-cell conformal intervals that
`impute()` reports. Those intervals summarise uncertainty about a single
imputed value and were close to 95% coverage in simulation (0.954 to 0.971,
GNN off, random missingness), but coverage can fall below 95% at small n
(about 100 species) and when missingness is clustered in clades.

`draws_method = "posterior"` fits a Bayesian multivariate phylogenetic mixed
model and samples each completed dataset from the posterior predictive
distribution of the missing cells. It currently requires:

- continuous traits only,
- one row per species (no `species_col`),
- no `covariates`.

For one incomplete covariate in a Gaussian `lm`, a binomial `glm` or an
`lmer` with one random intercept, `multi_impute_analysis()` provides a
narrower analysis-aware route; see `?multi_impute_analysis`.

## Impute

The bundled AVONET300 data are complete for the four continuous traits, so
we delete some values to have something to impute.

```{r impute}
library(pigauto)
data(avonet300, tree300)

traits <- avonet300[, c("Mass", "Wing.Length",
                        "Beak.Length_Culmen", "Tarsus.Length")]
rownames(traits) <- avonet300$Species_Key

set.seed(1)
traits$Mass[sample(nrow(traits), 60)] <- NA
traits$Wing.Length[sample(nrow(traits), 45)] <- NA

mi <- multi_impute(traits, tree300, m = 20L,
                   draws_method = "posterior", seed = 1L)
mi
```

With the default sampler settings, a simulated four-trait dataset of 300
species took about ten minutes per fit on one core (median 616 s, Totoro
server); fits that need automatic chain extensions take longer. Check
convergence before using the draws: `mi$posterior$diagnostics`
holds split R-hat and bulk effective sample size for each parameter, and
`multi_impute()` warns if the chains have not converged after its automatic
extensions.

```{r diagnostics}
mi$posterior$diagnostics
attr(mi$posterior$diagnostics, "converged")
```

`mi$datasets` is a list of `m` completed data frames with species as row
names. Observed values are unchanged in every dataset; only the missing
cells differ.

## Fit and pool

`with_imputations()` applies a model-fitting function to every completed
dataset and records where the datasets came from. `pool_mi()` then combines
the fixed-effect estimates with Rubin's rules (Rubin 1987; Barnard and Rubin
1999) and returns one row per coefficient with the pooled estimate, standard
error, degrees of freedom, confidence interval and the fraction of missing
information (`fmi`).

The question in every example is the same: how does wing length scale with
body mass?

### Phylogenetic regression (nlme)

```{r gls}
fits <- with_imputations(mi, function(d) {
  d$species <- rownames(d)
  nlme::gls(log(Wing.Length) ~ log(Mass), data = d,
            correlation = ape::corBrownian(phy = tree300, form = ~species),
            method = "ML")
})
pool_mi(fits)
```

### Mixed models (lme4 and glmmTMB)

Variables that were not imputed can be joined to each completed dataset
inside the fitting function. Here diet category (`Trophic.Level`) becomes a
random intercept.

```{r mixed}
add_diet <- function(d) {
  d$Trophic.Level <- avonet300$Trophic.Level[
    match(rownames(d), avonet300$Species_Key)]
  d
}

fits_lmer <- with_imputations(mi, function(d) {
  lme4::lmer(log(Wing.Length) ~ log(Mass) + (1 | Trophic.Level),
             data = add_diet(d))
})
pool_mi(fits_lmer)

fits_tmb <- with_imputations(mi, function(d) {
  glmmTMB::glmmTMB(log(Wing.Length) ~ log(Mass) + (1 | Trophic.Level),
                   data = add_diet(d))
})
pool_mi(fits_tmb)
```

For `glmmTMB` fits, `pool_mi()` pools the conditional (mean) model only;
zero-inflation and dispersion coefficients are not included.

A caveat on joined variables: the posterior imputation model conditions on
the imputed traits and the tree, not on `Trophic.Level`. When a variable
that matters for the analysis is left out of the imputation model, its
association with the imputed traits is weakened in the imputed cells.
Because the posterior route accepts continuous traits only, a categorical
variable such as diet cannot currently be added to the imputation model;
treat its estimated effects with this in mind. (Diet is also a small
grouping factor here, four levels, one of them with a single species, so
the random-intercept variance is poorly estimated.)

### Location-scale models (drmTMB)

`drmTMB` models the mean and the residual standard deviation together.
`pool_mi()` pools both sets of fixed effects, labelled `mu:` and `sigma:`.
The posterior imputation model has a constant residual variance, so the
imputed cells carry no relationship between body mass and residual spread.
We expect this to weaken a pooled `sigma:` slope when many cells are
imputed; this has not been measured.

```{r drmtmb}
fits_drm <- with_imputations(mi, function(d) {
  drmTMB::drmTMB(drmTMB::bf(log(Wing.Length) ~ log(Mass),
                            sigma ~ log(Mass)),
                 data = d)
})
pool_mi(fits_drm)
```

### Multivariate latent-variable models (gllvmTMB)

`gllvmTMB` fits all traits jointly in long format. `pool_mi()` pools its
fixed effects (here the trait means); latent loadings and variance
components are not pooled. gllvmTMB can be installed from GitHub with
`remotes::install_github("itchyshin/gllvmTMB")`.

```{r gllvmtmb}
to_long <- function(d) {
  tr <- names(d)
  data.frame(species = factor(rep(rownames(d), times = length(tr))),
             trait   = factor(rep(tr, each = nrow(d)), levels = tr),
             value   = log(unlist(d, use.names = FALSE)))
}

fits_gllvm <- with_imputations(mi, function(d) {
  gllvmTMB::gllvmTMB(value ~ 0 + trait + latent(0 + trait | species, d = 1),
                     data = to_long(d), trait = "trait", unit = "species",
                     silent = TRUE)
})
pool_mi(fits_gllvm)
```

### Other models

Any fit with `coef()` and `vcov()` methods can be pooled. For other classes,
pass extractor functions to `pool_mi()` through `coef_fun` and `vcov_fun`,
or a `tidy_fun` returning `term`, `estimate` and `std.error`.

## Bayesian models: concatenate, do not pool

Rubin's rules combine point estimates and standard errors. A Bayesian fit
already has a posterior distribution, and the right way to combine
imputations is to stack the posterior draws from all `m` fits into one
sample. `pool_mi()` refuses `brmsfit` and `MCMCglmm` objects for this
reason.

With brms, `brm_multiple()` fits every completed dataset and combines the
draws for you:

```{r brms}
fit_b <- brms::brm_multiple(log(Wing.Length) ~ log(Mass),
                            data = mi$datasets, chains = 2, refresh = 0)
summary(fit_b)
```

Check the per-dataset convergence diagnostics in `fit_b$rhats`; the combined
R-hat can look poor simply because the imputed datasets differ. Stacking
draws needs more imputations than Rubin's rules for stable results (Zhou
and Reiter 2010), so prefer a larger `m` than the 20 used here.

With MCMCglmm, fit each dataset and bind the fixed-effect samples:

```{r mcmcglmm}
fits_mc <- lapply(mi$datasets, function(d) {
  MCMCglmm::MCMCglmm(log(Wing.Length) ~ log(Mass), data = d, verbose = FALSE)
})
sol <- coda::as.mcmc(do.call(rbind, lapply(fits_mc, function(f) f$Sol)))
summary(sol)
```

## References

Barnard J, Rubin DB (1999). Small-sample degrees of freedom with multiple
imputation. *Biometrika* 86: 948-955.

Nakagawa S, Freckleton RP (2008). Missing inaction: the dangers of ignoring
missing data. *Trends in Ecology and Evolution* 23: 592-596.

Rubin DB (1987). *Multiple Imputation for Nonresponse in Surveys*. Wiley.

Zhou X, Reiter JP (2010). A note on Bayesian inference after multiple
imputation. *The American Statistician* 64: 159-163.
