---
title: "Propagating Tree Uncertainty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Propagating Tree Uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("torch", quietly = TRUE) &&
         isTRUE(try(torch::torch_is_installed(), silent = TRUE))
)
```

## Do I need this article?

**Short answer:** only if you have a **posterior sample of trees** (from BEAST,
MrBayes, BirdTree.org, etc.) and you want tree-topology uncertainty represented
in experimental Rubin-pooled fixed-effect standard errors and p-values.

Version 0.10.0 does not pool variance components, correlations, random-effect
predictions, BLUPs/conditional modes, or latent loadings. The MI workflow is
experimental pending known-DGP inferential coverage validation.

| Situation | Use this article? |
|---|---|
| One tree (published phylogeny, time-calibrated tree) | No — use `multi_impute()` and the [mixed-types vignette](mixed-types.html). |
| Posterior sample (2 or more trees) | Yes. |

## The two-step workflow

Tree uncertainty enters the analysis in two places. pigauto handles
step 1. Step 2 is your responsibility because the downstream model is
your choice.

```
+--------------------------------------+
| Step 1 -- imputation                 |
|                                      |
| multi_impute_trees(traits, trees)    |
|   -> T x m_per_tree completed        |
|      data.frames, each tagged with   |
|      the tree that produced it       |
+------------------+-------------------+
                   |
                   v
+--------------------------------------+
| Step 2 -- analysis + pool            |
|                                      |
| for dataset i:                       |
|   fit model with trees[[t_i]]        |
| pool_mi(fits)                        |
|                                      |
| The SAME tree that produced          |
| dataset i is used to fit model i.    |
+--------------------------------------+
```

## The canonical workflow (Nakagawa & de Villemereuil 2019)

With `share_gnn = TRUE` (the default), T = 50 posterior trees avoids fitting
50 separate GNNs. The literature workflow uses one imputation per tree
(M = 50 total), fits the downstream model 50 times (each with the matching
tree), and pools supported fixed effects with Rubin's rules.

```{r canonical, eval = FALSE}
library(pigauto)
data(avonet300, trees300)
df <- avonet300
rownames(df) <- df$Species_Key
df$Species_Key <- NULL

mi <- multi_impute_trees(df, trees = trees300, m_per_tree = 1L)
# share_gnn = TRUE, reference_tree = MCC via phangorn -- all default

fits <- with_imputations(mi, function(dat, tree) {
  dat$species <- rownames(dat)
  nlme::gls(
    log(Mass) ~ log(Wing.Length),
    correlation = ape::corBrownian(phy = tree, form = ~species),
    data = dat, method = "ML"
  )
})
pool_mi(fits)   # experimental fixed-effect pooling across both uncertainties
```

The code above is illustrative — full execution takes ~25 min because
it fits pigauto on the MCC reference tree, then runs a GLS model for
each of the 50 posterior trees. Running the chunk is left to the reader.

## Why `share_gnn = TRUE` preserves tree signal

The calibrated gate `r_cal` controls how much of each prediction comes
from the baseline vs the GNN. In high-phylogenetic-signal regimes the
gate often closes or nearly closes, so `pred = baseline(tree_t)` and the
per-tree baseline carries the tree-uncertainty signal. When the gate is
partly open, the GNN component is shared across trees and the per-tree
baseline still varies with `tree_t`. See `?multi_impute_trees` under
"Share-GNN (tree-sharing) mode" for the fully-open and partially-open
cases.

If you need exact per-tree model independence (e.g. for methodological
comparison), set `share_gnn = FALSE`:

```{r opt_out, eval = FALSE}
mi_slow <- multi_impute_trees(df, trees300, m_per_tree = 1L,
                               share_gnn = FALSE)
# fits T = length(trees300) full pigauto models -- ~10-15x slower.
```

## Scale choices

| T | m_per_tree | M | When |
|---|---|---|---|
| 50 | 1 | 50 | Literature example following N&dV 2019. |
| 20 | 2 | 40 | Smaller posterior; more than one draw per tree. |
| 10 | 5 | 50 | Very small posterior; per-tree variance helps. |
| <10 | bump m_per_tree | >=25 | Runtime warning fires; Rubin's rules unstable below M=25. |

## References

- Nakagawa S, de Villemereuil P (2019). A general method for
  simultaneously accounting for phylogenetic and species sampling
  uncertainty via Rubin's rules in comparative analysis.
  *Systematic Biology* 68(4): 632-641.
- Rubin DB (1987). *Multiple Imputation for Nonresponse in Surveys*.
  Wiley.
- Jetz W et al. (2012). The global diversity of birds in space and
  time. *Nature* 491: 444-448.
