---
title: "Frequentist and Bayesian multiple imputation of phylogenetic traits under Rubin's rules"
subtitle: "Simulation study record: design, compute and results"
date: last-modified
format:
  html:
    toc: true
    embed-resources: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
d <- "data"; ag <- file.path(d, "agg")
rd <- function(f) utils::read.csv(file.path(ag, f), stringsAsFactors = FALSE)
cells_l <- rd("cells_l.csv"); cells_n <- rd("cells_n.csv"); down_l <- rd("down_l.csv"); down_n <- rd("down_n.csv")
paired <- rd("paired.csv"); failures <- rd("failures.csv")
fits <- utils::read.csv(file.path(d, "fits.csv"), stringsAsFactors = FALSE)
diag <- utils::read.csv(file.path(d, "bace_diagnostics.csv"), stringsAsFactors = FALSE)
disc_dir <- file.path(d, "discrete"); has_disc <- dir.exists(disc_dir) && length(list.files(disc_dir, "\\.csv$")) > 0
lab <- c(complete = "complete data", freqA = "freq A (proper)", freqB = "freq B (improper)",
         bace = "BACE as shipped", bace_chain = "chained BACE", bace_resid = "BACE + residual (control)")
ord <- names(lab)
f3 <- function(x) formatC(x, format = "f", digits = 3)
tab <- function(x, caption) knitr::kable(x, caption = caption, align = "l", row.names = FALSE)
wide <- function(df, value, se = NULL, rows = c("n", "lambda")) {
  df <- df[df$arm %in% ord, ]
  df$cell <- if (is.null(se)) f3(df[[value]]) else paste0(f3(df[[value]]), " (", f3(df[[se]]), ")")
  w <- stats::reshape(df[, c(rows, "arm", "cell")], idvar = rows, timevar = "arm", direction = "wide")
  names(w) <- sub("^cell\\.", "", names(w))
  keep <- intersect(ord, names(w)); w <- w[, c(rows, keep)]; names(w)[-seq_along(rows)] <- lab[keep]
  w[do.call(order, w[rows]), ]
}
```

# Summary

We compared four ways of multiply imputing missing phylogenetic trait values, plus a negative control, on the
same simulated datasets, and scored each with Rubin's rules: per missing value and on two downstream estimates.
The continuous part is complete: `r format(sum(fits$set == "bace"), big.mark = ",")` BACE fits and
`r format(sum(fits$set == "freq"), big.mark = ",")` frequentist fits over 18 simulation cells, with no dataset
missing. The discrete part (binary, ordinal, categorical traits) extends the same design on the same datasets
(@sec-discrete).

Main continuous findings:

- **BACE as shipped gives improper imputations.** Its 20 final runs all start from the same converged dataset,
  so the imputations are not independent draws. Per-value coverage is below nominal at λ = 0.3 and 0.7 (about
  0.91), and downstream coverage of the phylogenetic correlation is about 0.90 at n = 300 and 1000.
- **Chaining BACE's final runs fixes coverage.** Starting final run *m* from dataset *m* − 1 gives near-nominal
  per-value coverage; downstream it is slightly conservative.
- **Proper frequentist imputation (freq A) is close to nominal throughout**; the improper version (freq B) is
  fine per value but under-covers the correlation, so improperness shows up downstream.
- **Accuracy.** freq A and chained BACE are equally accurate at n = 300 and 1000; at n = 100 chained BACE is less
  accurate and pulls the downstream slope toward zero.

This is the frequentist method's home ground: the scored traits are continuous, evolve by Brownian motion with
Pagel's λ, and Rphylopars fits exactly that model. It does not test BACE's claimed advantage, one model for
mixed trait types; the discrete extension does.

# Design (ADEMP)

## Aims

Does BACE, a Bayesian chained-equation imputer built on MCMCglmm, give proper multiple imputations, measured by
Rubin-pooled coverage and bias against a proper and an improper frequentist imputer, when the frequentist model
is correctly specified?

## Data-generating mechanism

- Tree: `ape::rcoal(n)`; tip correlation matrix *C* = cov2cor(vcv(tree)).
- Latent liabilities for seven traits: vec(*L*) ~ N(0, Σ~ρ~ ⊗ *V*~λ~), with *V*~λ~ = λ*C* + (1 − λ)*I* (Pagel's λ
  on the correlation scale) and Σ~ρ~ with unit variances and every off-diagonal equal to ρ.
- Observed traits: two continuous (c1, c2), a count, a proportion (prp), a binary trait, a four-level ordinal
  trait and a three-class categorical trait, derived from the liabilities with thresholds fixed at population
  quantiles, plus a fully observed driver (d1). Exact transforms: `make_cell()` in
  `script/campaign_gnn_off_lib.R`.
- Missingness: MCAR, 30% of cells of the seven traits (realised fraction
  `r f3(mean(fits$realised_frac))` on average); the driver is always observed.
- Factors (full factorial, 18 cells): n ∈ {100, 300, 1000}, λ ∈ {0.3, 0.7, 1}, ρ ∈ {0, 0.5}.
- Replicates: 200 datasets per cell for every arm, except BACE at n = 1000 with 100 datasets per cell. Datasets
  are indexed by seed; each arm set regenerates the same dataset from (n, λ, ρ, seed), confirmed by identical
  complete-data estimates across arm sets.

Under this mechanism c2 = ρ·c1 + *e* with *e* ~ N(0, (1 − ρ²)*V*~λ~), so the population PGLS slope of c2 on c1
is ρ, the model `nlme::gls(c2 ~ c1, corPagel(λ))` is correctly specified, and the phylogenetic correlation of
c1 and c2 is also ρ.

## Estimands

1. **Each missing value** of c1 and c2 (and prp, on the logit scale, reported separately).
2. **PGLS slope** of c2 on c1 (Pagel's λ estimated by REML, bounded to [0, 1]); true value ρ.
3. **Phylogenetic correlation** of c1 and c2 (GLS-whitened with the estimated λ, Fisher's z, variance
   1/(n − 3), complete-data df treated as infinite); true value ρ.

## Methods (arms), M = 20 imputations each

| arm | how the 20 imputations are made | role |
|---|---|---|
| freq A | Rphylopars, Pagel's λ, one joint fit of c1, c2, logit(prp) and the driver; for each *m*, a parametric bootstrap refit, then a joint conditional draw of the missing cells under the refit parameters | proper frequentist MI |
| freq B | the same model fitted once; 20 joint conditional draws with parameters held fixed | improper contrast |
| BACE as shipped | BACE (MCMCglmm chained equations), runs 5, nitt 50,000, burnin 10,000, thin 25; its 20 final datasets | what a BACE user gets |
| chained BACE | BACE's own final step run 20 times, each run starting from the previous draw; same fitted chains | BACE with independent draws |
| BACE + residual | BACE as shipped plus an extra residual draw (σ² from the posterior) | negative control |
| complete data | the same analyses on the data before masking | reference |

Frequentist safeguards: Cholesky solves, hard bounds on every conditional distribution, and a plausibility
check on every bootstrap refit (implied tip variance within a factor of 50 of the original fit); a failed draw
redraws the bootstrap sample and is counted. BACE inputs are cleaned before fitting: unused factor levels are
dropped and a discrete trait with fewer than two observed classes is left out (otherwise BACE stops at λ = 1).

## Performance measures

- Per value: coverage of the Rubin interval mean ± *t*~M−1~·√((1 + 1/M)·*B*) (within-imputation variance is zero
  for a missing value), interval width, z-RMSE (RMSE of the pooled mean divided by the trait's SD) and interval
  score. Coverage is averaged within a dataset over c1 and c2, then over datasets.
- Downstream: Rubin pooling with Barnard–Rubin degrees of freedom; coverage of ρ, bias, RMSE, interval width,
  fraction of missing information.
- Monte Carlo SEs are over datasets (clustered by dataset). A 20-draw percentile interval would cover only 0.872
  under a correct model, so no percentile intervals are used.

# Computing

```{r compute}
fits$hours <- fits$wall_s_total / 3600
host <- as.data.frame.matrix(table(fits$cluster, fits$set)); host$nodes <- tapply(fits$node, fits$cluster, function(x) length(unique(x)))[rownames(host)]
host <- cbind(cluster = rownames(host), host); rownames(host) <- NULL
tab(host, "Fits of record by cluster (after removing duplicate submissions) and number of distinct compute nodes.")
w <- aggregate(hours ~ set + n, fits, function(x) c(median = median(x), p90 = unname(quantile(x, 0.9)), max = max(x)))
w <- data.frame(set = w$set, n = w$n, round(w$hours, 2)); names(w)[3:5] <- c("median h", "90th pct h", "max h")
tab(w[order(w$set, w$n), ], "Wall time per fit (one dataset, all arms of the set), single-threaded.")
```

| host | hardware | used for |
|---|---|---|
| fir (Digital Research Alliance of Canada) | 192-core nodes, 768 GB; Slurm, 1-day limit | frequentist campaign (all 3,600); BACE at n = 300 and 1000 |
| rorqual (Alliance) | 192-core nodes, 768 GB; Slurm, 3-day limit | BACE at n = 100, 300, 1000 |
| nibi (Alliance) | 192-core nodes, 766 GB; Slurm, 3-day limit | BACE at n = 100, 300, 1000; BACE settings pre-run (240 fits) |
| Totoro (University of Alberta) | AMD EPYC 9655, 384 cores, about 1 TB; no queue; capped at 150 cores | BACE at n = 1000 (70 fits) |

Every fit ran on one core (`OPENBLAS_NUM_THREADS = 1`). Total single-core time of the fits of record: about
`r format(round(sum(fits$hours[fits$set == "bace"])), big.mark = ",")` hours for BACE and
`r format(round(sum(fits$hours[fits$set == "freq"])), big.mark = ",")` for the frequentist arms. Slurm
accounting over all campaign jobs, including reruns, duplicates and superseded passes, was about 6,750
core-hours (fir 2,350, rorqual 2,231, nibi 2,169). Peak memory for one BACE fit at n = 1000 was 18.8 GB.
Runs: `r min(substr(fits$time, 1, 10))` to `r max(substr(fits$time, 1, 10))`.

```{r versions}
v <- function(x) paste(names(table(x)), collapse = ", ")
tab(data.frame(software = c("R", "BACE", "MCMCglmm", "Rphylopars"),
               version = c(v(fits$R), v(fits$BACE), v(fits$MCMCglmm), v(fits$Rphylopars))),
    "Software versions recorded in the result files.")
```

BACE reports version 0.1.0 on the Alliance clusters and 0.0.0.9000 on Totoro; its fitting functions were checked
byte-identical across all hosts. The cluster copies of the scripts are not git checkouts, so the git hash is not
recorded per fit; the code of record is branch `arc/rubin-freq-bace` of pigauto.

# Results: continuous traits

## Per-value coverage

```{r cellcov}
tab(wide(cells_l, "coverage", "coverage_se"), "Per-value coverage of the Rubin interval for missing c1 and c2 (Monte Carlo SE), pooled over ρ. Nominal 0.95.")
```

```{r fig-cov, fig.width = 9, fig.height = 5.5, fig.cap = "Per-value coverage by sample size and λ; whiskers ±2 Monte Carlo SE."}
arms <- c("freqA", "freqB", "bace", "bace_chain", "bace_resid"); cols <- c("#e69f00", "#cc79a7", "#d55e00", "#0072b2", "#009e73")
op <- par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
for (l in c(0.3, 0.7, 1)) {
  s <- cells_l[cells_l$lambda == l & cells_l$arm %in% arms, ]
  plot(NA, xlim = c(0.7, 3.3), ylim = c(0.85, 1), xaxt = "n", xlab = "n", ylab = "coverage", main = paste("λ =", l))
  axis(1, 1:3, c(100, 300, 1000)); abline(h = 0.95, lty = 2, col = "grey40")
  for (j in seq_along(arms)) { r <- s[s$arm == arms[j], ]; r <- r[order(r$n), ]; x <- match(r$n, c(100, 300, 1000)) + (j - 3) * 0.08
    segments(x, r$coverage - 2 * r$coverage_se, x, r$coverage + 2 * r$coverage_se, col = cols[j]); points(x, r$coverage, pch = 19, col = cols[j]) }
  if (l == 0.3) legend("bottomright", lab[arms], col = cols, pch = 19, bty = "n", cex = 0.8)
}
par(op)
```

## Downstream estimates

```{r down, results = "asis"}
for (e in c("slope", "cor")) {
  s <- down_n[down_n$estimand == e & down_n$arm %in% ord, ]
  s <- s[order(s$n, match(s$arm, ord)), ]
  cat(tab(data.frame(n = s$n, arm = lab[s$arm], coverage = paste0(f3(s$coverage), " (", f3(s$coverage_se), ")"),
                       bias = paste0(f3(s$bias), " (", f3(s$bias_se), ")"), RMSE = f3(s$rmse), `CI width` = f3(s$ci_width),
                       FMI = f3(s$fmi_median), datasets = s$datasets, check.names = FALSE),
            paste0("Rubin-pooled ", if (e == "slope") "PGLS slope of c2 on c1" else "phylogenetic correlation of c1 and c2",
                   ": coverage of ρ and bias (Monte Carlo SE), pooled over λ and ρ.")), sep = "\n"); cat("\n\n")
}
```

```{r down-l}
tab(wide(down_l[down_l$estimand == "cor", ], "coverage"), "Coverage of ρ by the pooled phylogenetic correlation, by n and λ.")
```

## Accuracy

```{r acc}
tab(wide(cells_l, "zrmse", "zrmse_se"), "z-RMSE of the pooled imputation for missing c1 and c2 (SD units; Monte Carlo SE). Lower is better.")
tab(wide(cells_l, "iscore"), "Interval score of the per-value Rubin interval. Lower is better.")
```

## Paired contrasts

```{r paired}
p <- paired; p$diff <- paste0(f3(p$diff), " (", f3(p$se), ")")
tab(p[, c("contrast", "measure", "n", "lambda", "diff", "datasets")], "Paired differences on the same datasets (SE over datasets).")
```

## Failures and diagnostics

```{r fail}
tab(failures, "Per n and λ: result files, BACE and Rphylopars errors inside a file, BACE fits needing input cleaning, and bootstrap refits redrawn by the frequentist safeguards.")
```

- `r nrow(utils::read.csv(file.path(d, "failure_records.csv")))` first-pass BACE fits failed before input cleaning
  (an unobserved discrete level at λ = 1) and are kept as records in `data/failure_records.csv`; each was rerun
  with cleaning.
- BACE fits that still errored inside a result file are recorded but not scored, so BACE rows rest on fewer
  datasets than the design (see the `datasets` columns).
- BACE's own convergence check passed on `r f3(mean(diag$converged, na.rm = TRUE))` of fits; median ESS `r round(median(diag$ess_med, na.rm = TRUE))`.
  The check tests BACE's deterministic convergence phase and is
  scale-dependent, so it was not used to select settings (pre-run page, 240 fits).
- MCMCglmm segmentation faults and scheduler timeouts left no result file; every affected dataset was rerun
  until the completeness check showed none missing.

# Discrete traits {#sec-discrete}

```{r disc, results = "asis"}
if (!has_disc) cat("*Not yet run.* Results will appear here from `data/discrete/`.\n")
```

The discrete extension mirrors the continuous design:

| element | continuous (done) | discrete (planned) |
|---|---|---|
| datasets | 18 cells, seeds as above | the **same** cells, seeds and datasets (`bin`, `ord`, `cat3` of each dataset) |
| M | 20 | 20 |
| proper frequentist | freq A: bootstrap refit, then draw | castor Mk hidden-state prediction with a parametric bootstrap of the rates before each draw |
| improper frequentist | freq B: one fit | castor Mk with rates held at the fitted values (as in simulation v1) |
| Bayesian | BACE as shipped; chained BACE | the same BACE fits (re-run with imputations saved; the continuous rows should reproduce) |
| per-value scores | coverage, width, z-RMSE, interval score | class accuracy, Brier score, calibration, coverage of the 20-draw prediction set |
| downstream, Rubin-pooled | PGLS slope, phylogenetic correlation | an estimate involving a discrete trait (for example the slope of c1 on `bin`), truth computed on the liability scale or by Monte Carlo |
| replicates | 200 per cell (BACE 100 at n = 1000) | n = 100 and 300 first, same counts; n = 1000 decided after |

castor models one trait at a time, while BACE conditions each trait on the others, so the contrast tests
whether BACE's use of correlated traits pays off. Simulation v1 (accuracy only, earlier BACE settings) found
BACE ahead at λ = 0.3 and castor ahead at λ = 1.

# Limitations

- One mechanism: Brownian motion with Pagel's λ, MCAR 30%, random coalescent trees. No MAR or clade-structured
  missingness, no OU, no real trees.
- The frequentist model shares one λ across its traits; prp carries extra non-phylogenetic noise, so its fitted
  variance is inflated at λ near 1 (c1 and c2 are unaffected).
- BACE at n = 1000 has 100 datasets per cell, so its Monte Carlo SEs are about 1.4 times the others'.
- At λ = 1 and n = 1000 freq A and chained BACE both cover slightly above 0.95; not investigated.

# Files and reproduction

| file | contents |
|---|---|
| `data/fits.csv` | one row per fit: cell, seed, cluster, node, time, wall times, errors, versions |
| `data/fit_cells.csv.gz` | fit × arm × trait: per-value coverage, width, z-RMSE, interval score |
| `data/fit_estimands.csv.gz` | fit × arm × estimand: pooled estimate, SE, interval, df, FMI, truth, covered |
| `data/bace_diagnostics.csv` | BACE convergence verdict, drift, ESS, input cleaning |
| `data/failure_records.csv` | first-pass BACE failures kept as records |
| `data/agg/` | aggregated tables used above, and `report.json` for the web pages |
| `data/discrete/` | discrete results (to come) |

Pipeline (worktree root): `script/rubin_cell.R` (one dataset, all arms of a set) run on the clusters with
`script/rubin_campaign_nibi.sh` and `script/rubin_totoro.sh`; `script/rubin_pool.sh` copies results to the pool;
`script/rubin_missing.R` checks completeness; `script/rubin_campaign_aggregate.R` aggregates;
`script/rubin_study/build_data.R` writes this folder's tables; `quarto render script/rubin_study/study.qmd`.
Tests: `script/tests-rubin/`. The raw per-fit files (556 MB) stay in the pool and on the clusters.
