---
title: "Frequentist and Bayesian multiple imputation of phylogenetic traits under Rubin's rules"
subtitle: "Simulation study record: design, compute and results"
date: last-modified
format:
  html:
    toc: true
    embed-resources: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
d <- "data"; ag <- file.path(d, "agg")
rd <- function(f) utils::read.csv(file.path(ag, f), stringsAsFactors = FALSE)
cells_l <- rd("cells_l.csv"); cells_n <- rd("cells_n.csv"); down_l <- rd("down_l.csv"); down_n <- rd("down_n.csv")
paired <- rd("paired.csv"); failures <- rd("failures.csv")
fits <- utils::read.csv(file.path(d, "fits.csv"), stringsAsFactors = FALSE)
diag <- utils::read.csv(file.path(d, "bace_diagnostics.csv"), stringsAsFactors = FALSE)
disc_dir <- file.path(d, "discrete")   # the truth table lives there before any result, so test for the aggregates
has_disc <- all(file.exists(file.path(disc_dir, c("agg_disc_cells.csv", "agg_disc_cells_l.csv", "agg_disc_down.csv", "agg_disc_paired.csv"))))
lab <- c(complete = "complete data", freqA = "freq A (proper)", freqB = "freq B (improper)",
         bace = "BACE as shipped", bace_chain = "chained BACE", bace_resid = "BACE + residual (control)")
ord <- names(lab)
f3 <- function(x) formatC(x, format = "f", digits = 3)
tab <- function(x, caption) knitr::kable(x, caption = caption, align = "l", row.names = FALSE)
wide <- function(df, value, se = NULL, rows = c("n", "lambda")) {
  df <- df[df$arm %in% ord, ]
  df$cell <- if (is.null(se)) f3(df[[value]]) else paste0(f3(df[[value]]), " (", f3(df[[se]]), ")")
  w <- stats::reshape(df[, c(rows, "arm", "cell")], idvar = rows, timevar = "arm", direction = "wide")
  names(w) <- sub("^cell\\.", "", names(w))
  keep <- intersect(ord, names(w)); w <- w[, c(rows, keep)]; names(w)[-seq_along(rows)] <- lab[keep]
  w[do.call(order, w[rows]), ]
}
```

# Summary

We compared four ways of multiply imputing missing phylogenetic trait values, plus a negative control, on the
same simulated datasets, and scored each with Rubin's rules: per missing value and on two downstream estimates.
The continuous part is complete: `r format(sum(fits$set == "bace"), big.mark = ",")` BACE fits and
`r format(sum(fits$set == "freq"), big.mark = ",")` frequentist fits over 18 simulation cells, with no dataset
missing. The discrete part (binary, ordinal, categorical traits) extends the same design on the same datasets
(@sec-discrete).

Main continuous findings:

- **BACE as shipped gives improper imputations.** Its 20 final runs all start from the same converged dataset,
  so the imputations are not independent draws. Per-value coverage is below nominal at λ = 0.3 and 0.7 (about
  0.91), and downstream coverage of the phylogenetic correlation is about 0.90 at n = 300 and 1000.
- **Chaining BACE's final runs fixes coverage.** Starting final run *m* from dataset *m* − 1 gives near-nominal
  per-value coverage; downstream it is slightly conservative.
- **Proper frequentist imputation (freq A) is close to nominal throughout**; the improper version (freq B) is
  fine per value but under-covers the correlation, so improperness shows up downstream.
- **Accuracy.** freq A and chained BACE are equally accurate at n = 300 and 1000; at n = 100 chained BACE is less
  accurate and pulls the downstream slope toward zero.

This is the frequentist method's home ground: the scored traits are continuous, evolve by Brownian motion with
Pagel's λ, and Rphylopars fits exactly that model. It does not test BACE's claimed advantage, one model for
mixed trait types; the discrete extension does.

# Design (ADEMP)

## Aims

Does BACE, a Bayesian chained-equation imputer built on MCMCglmm, give proper multiple imputations, measured by
Rubin-pooled coverage and bias against a proper and an improper frequentist imputer, when the frequentist model
is correctly specified?

## Data-generating mechanism

- Tree: `ape::rcoal(n)`; tip correlation matrix *C* = cov2cor(vcv(tree)).
- Latent liabilities for seven traits: vec(*L*) ~ N(0, Σ~ρ~ ⊗ *V*~λ~), with *V*~λ~ = λ*C* + (1 − λ)*I* (Pagel's λ
  on the correlation scale) and Σ~ρ~ with unit variances and every off-diagonal equal to ρ.
- Observed traits: two continuous (c1, c2), a count, a proportion (prp), a binary trait, a four-level ordinal
  trait and a three-class categorical trait, derived from the liabilities with thresholds fixed at population
  quantiles, plus a fully observed driver (d1). Exact transforms: `make_cell()` in
  `script/campaign_gnn_off_lib.R`.
- Missingness: MCAR, 30% of cells of the seven traits (realised fraction
  `r f3(mean(fits$realised_frac))` on average); the driver is always observed.
- Factors (full factorial, 18 cells): n ∈ {100, 300, 1000}, λ ∈ {0.3, 0.7, 1}, ρ ∈ {0, 0.5}.
- Replicates: 200 datasets per cell for every arm, except BACE at n = 1000 with 100 datasets per cell. Datasets
  are indexed by seed; each arm set regenerates the same dataset from (n, λ, ρ, seed), confirmed by identical
  complete-data estimates across arm sets.

Under this mechanism c2 = ρ·c1 + *e* with *e* ~ N(0, (1 − ρ²)*V*~λ~), so the population PGLS slope of c2 on c1
is ρ, the model `nlme::gls(c2 ~ c1, corPagel(λ))` is correctly specified, and the phylogenetic correlation of
c1 and c2 is also ρ.

## Estimands

1. **Each missing value** of c1 and c2 (and prp, on the logit scale, reported separately).
2. **PGLS slope** of c2 on c1 (Pagel's λ estimated by REML, bounded to [0, 1]); true value ρ.
3. **Phylogenetic correlation** of c1 and c2 (GLS-whitened with the estimated λ, Fisher's z, variance
   1/(n − 3), complete-data df treated as infinite); true value ρ.

## Methods (arms), M = 20 imputations each

| arm | how the 20 imputations are made | role |
|---|---|---|
| freq A | Rphylopars, Pagel's λ, one joint fit of c1, c2, logit(prp) and the driver; for each *m*, a parametric bootstrap refit, then a joint conditional draw of the missing cells under the refit parameters | proper frequentist MI |
| freq B | the same model fitted once; 20 joint conditional draws with parameters held fixed | improper contrast |
| BACE as shipped | BACE (MCMCglmm chained equations), runs 5, nitt 50,000, burnin 10,000, thin 25; its 20 final datasets | what a BACE user gets |
| chained BACE | BACE's own final step run 20 times, each run starting from the previous draw; same fitted chains | BACE with independent draws |
| BACE + residual | BACE as shipped plus an extra residual draw (σ² from the posterior) | negative control |
| complete data | the same analyses on the data before masking | reference |

Frequentist safeguards: Cholesky solves, hard bounds on every conditional distribution, and a plausibility
check on every bootstrap refit (implied tip variance within a factor of 50 of the original fit); a failed draw
redraws the bootstrap sample and is counted. BACE inputs are cleaned before fitting: unused factor levels are
dropped and a discrete trait with fewer than two observed classes is left out (otherwise BACE stops at λ = 1).

## Performance measures

- Per value: coverage of the Rubin interval mean ± *t*~M−1~·√((1 + 1/M)·*B*) (within-imputation variance is zero
  for a missing value), interval width, z-RMSE (RMSE of the pooled mean divided by the trait's SD) and interval
  score. Coverage is averaged within a dataset over c1 and c2, then over datasets.
- Downstream: Rubin pooling with Barnard–Rubin degrees of freedom; coverage of ρ, bias, RMSE, interval width,
  fraction of missing information.
- Monte Carlo SEs are over datasets (clustered by dataset). A 20-draw percentile interval would cover only 0.872
  under a correct model, so no percentile intervals are used.

# Computing

```{r compute}
fits$hours <- fits$wall_s_total / 3600
host <- as.data.frame.matrix(table(fits$cluster, fits$set)); host$nodes <- tapply(fits$node, fits$cluster, function(x) length(unique(x)))[rownames(host)]
host <- cbind(cluster = rownames(host), host); rownames(host) <- NULL
tab(host, "Fits of record by cluster (after removing duplicate submissions) and number of distinct compute nodes.")
w <- aggregate(hours ~ set + n, fits, function(x) c(median = median(x), p90 = unname(quantile(x, 0.9)), max = max(x)))
w <- data.frame(set = w$set, n = w$n, round(w$hours, 2)); names(w)[3:5] <- c("median h", "90th pct h", "max h")
tab(w[order(w$set, w$n), ], "Wall time per fit (one dataset, all arms of the set), single-threaded.")
```

| host | hardware | used for |
|---|---|---|
| fir (Digital Research Alliance of Canada) | 192-core nodes, 768 GB; Slurm, 1-day limit | frequentist campaign (all 3,600); BACE at n = 300 and 1000 |
| rorqual (Alliance) | 192-core nodes, 768 GB; Slurm, 3-day limit | BACE at n = 100, 300, 1000 |
| nibi (Alliance) | 192-core nodes, 766 GB; Slurm, 3-day limit | BACE at n = 100, 300, 1000; BACE settings pre-run (240 fits) |
| Totoro (University of Alberta) | AMD EPYC 9655, 384 cores, about 1 TB; no queue; capped at 150 cores | BACE at n = 1000 (70 fits) |

Every fit ran on one core (`OPENBLAS_NUM_THREADS = 1`). Total single-core time of the fits of record: about
`r format(round(sum(fits$hours[fits$set == "bace"])), big.mark = ",")` hours for BACE and
`r format(round(sum(fits$hours[fits$set == "freq"])), big.mark = ",")` for the frequentist arms. Slurm
accounting over all campaign jobs, including reruns, duplicates and superseded passes, was about 6,750
core-hours (fir 2,350, rorqual 2,231, nibi 2,169). Peak memory for one BACE fit at n = 1000 was 18.8 GB.
Runs: `r min(substr(fits$time, 1, 10))` to `r max(substr(fits$time, 1, 10))`.

```{r versions}
v <- function(x) paste(names(table(x)), collapse = ", ")
tab(data.frame(software = c("R", "BACE", "MCMCglmm", "Rphylopars"),
               version = c(v(fits$R), v(fits$BACE), v(fits$MCMCglmm), v(fits$Rphylopars))),
    "Software versions recorded in the result files.")
```

BACE reports version 0.1.0 on the Alliance clusters and 0.0.0.9000 on Totoro; its fitting functions were checked
byte-identical across all hosts. The cluster copies of the scripts are not git checkouts, so the git hash is not
recorded per fit; the code of record is branch `arc/rubin-freq-bace` of pigauto.

# Results: continuous traits

## Per-value coverage

```{r cellcov}
tab(wide(cells_l, "coverage", "coverage_se"), "Per-value coverage of the Rubin interval for missing c1 and c2 (Monte Carlo SE), pooled over ρ. Nominal 0.95.")
```

```{r fig-cov, fig.width = 9, fig.height = 5.5, fig.cap = "Per-value coverage by sample size and λ; whiskers ±2 Monte Carlo SE."}
arms <- c("freqA", "freqB", "bace", "bace_chain", "bace_resid"); cols <- c("#e69f00", "#cc79a7", "#d55e00", "#0072b2", "#009e73")
op <- par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
for (l in c(0.3, 0.7, 1)) {
  s <- cells_l[cells_l$lambda == l & cells_l$arm %in% arms, ]
  plot(NA, xlim = c(0.7, 3.3), ylim = c(0.85, 1), xaxt = "n", xlab = "n", ylab = "coverage", main = paste("λ =", l))
  axis(1, 1:3, c(100, 300, 1000)); abline(h = 0.95, lty = 2, col = "grey40")
  for (j in seq_along(arms)) { r <- s[s$arm == arms[j], ]; r <- r[order(r$n), ]; x <- match(r$n, c(100, 300, 1000)) + (j - 3) * 0.08
    segments(x, r$coverage - 2 * r$coverage_se, x, r$coverage + 2 * r$coverage_se, col = cols[j]); points(x, r$coverage, pch = 19, col = cols[j]) }
  if (l == 0.3) legend("bottomright", lab[arms], col = cols, pch = 19, bty = "n", cex = 0.8)
}
par(op)
```

## Downstream estimates

```{r down, results = "asis"}
for (e in c("slope", "cor")) {
  s <- down_n[down_n$estimand == e & down_n$arm %in% ord, ]
  s <- s[order(s$n, match(s$arm, ord)), ]
  cat(tab(data.frame(n = s$n, arm = lab[s$arm], coverage = paste0(f3(s$coverage), " (", f3(s$coverage_se), ")"),
                       bias = paste0(f3(s$bias), " (", f3(s$bias_se), ")"), RMSE = f3(s$rmse), `CI width` = f3(s$ci_width),
                       FMI = f3(s$fmi_median), datasets = s$datasets, check.names = FALSE),
            paste0("Rubin-pooled ", if (e == "slope") "PGLS slope of c2 on c1" else "phylogenetic correlation of c1 and c2",
                   ": coverage of ρ and bias (Monte Carlo SE), pooled over λ and ρ.")), sep = "\n"); cat("\n\n")
}
```

```{r down-l}
tab(wide(down_l[down_l$estimand == "cor", ], "coverage"), "Coverage of ρ by the pooled phylogenetic correlation, by n and λ.")
```

## Accuracy

```{r acc}
tab(wide(cells_l, "zrmse", "zrmse_se"), "z-RMSE of the pooled imputation for missing c1 and c2 (SD units; Monte Carlo SE). Lower is better.")
tab(wide(cells_l, "iscore"), "Interval score of the per-value Rubin interval. Lower is better.")
```

## Paired contrasts

```{r paired}
p <- paired; p$diff <- paste0(f3(p$diff), " (", f3(p$se), ")")
tab(p[, c("contrast", "measure", "n", "lambda", "diff", "datasets")], "Paired differences on the same datasets (SE over datasets).")
```

## Failures and diagnostics

```{r fail}
tab(failures, "Per n and λ: result files, BACE and Rphylopars errors inside a file, BACE fits needing input cleaning, and bootstrap refits redrawn by the frequentist safeguards.")
```

- `r nrow(utils::read.csv(file.path(d, "failure_records.csv")))` first-pass BACE fits failed before input cleaning
  (an unobserved discrete level at λ = 1) and are kept as records in `data/failure_records.csv`; each was rerun
  with cleaning.
- BACE fits that still errored inside a result file are recorded but not scored, so BACE rows rest on fewer
  datasets than the design (see the `datasets` columns).
- BACE's own convergence check passed on `r f3(mean(diag$converged, na.rm = TRUE))` of fits; median ESS `r round(median(diag$ess_med, na.rm = TRUE))`.
  The check tests BACE's deterministic convergence phase and is
  scale-dependent, so it was not used to select settings (pre-run page, 240 fits).
- MCMCglmm segmentation faults and scheduler timeouts left no result file; every affected dataset was rerun
  until the completeness check showed none missing.

# Discrete traits {#sec-discrete}

```{r disc, results = "asis"}
if (!has_disc) cat("*Not yet run.* Results will appear here from `data/discrete/`.\n")
```

The discrete extension mirrors the continuous design:

| element | continuous | discrete |
|---|---|---|
| datasets | 18 cells, seeds as above | the **same** cells, seeds and datasets (`bin`, `ord`, `cat3` of each dataset) |
| M | 20 | 20 |
| proper frequentist | freq A: bootstrap refit, then draw | castor Mk fit, then for each *m* a parametric bootstrap refit of the rates and a joint conditional draw of the missing tips (forward filtering, backward sampling on the tree); filled into freq A's continuous draws |
| improper frequentist | freq B: one fit | the same castor fit, 20 joint conditional draws at the fitted rates; filled into freq B's continuous draws |
| rate model | Pagel's λ (falls back to the sample mean at low signal) | all rates different (ARD) for `bin` and `cat3`, stepwise with every rate different (SRD) for `ord`: their stationary distribution can match the observed class frequencies, so they fall back to those at low signal |
| comparison arms | none | freq A and freq B with simulation v1's equal-rates models (ER, ER, SUEDE), scored on discrete traits only |
| Bayesian | BACE as shipped; chained BACE | the same BACE fits, re-run with the imputations saved |
| per-value scores | coverage, width, z-RMSE, interval score | class accuracy, Brier score, coverage and size of the 95% class set from the 20 draws |
| downstream, Rubin-pooled | PGLS slope, phylogenetic correlation | PGLS slope of c1 on `bin`, scored against a per-dataset target and a population target (below) |
| replicates | 200 per cell (BACE 100 at n = 1000) | n = 100 and 300 first, same counts; n = 1000 decided after |

castor models one trait at a time, while BACE conditions each trait on the others, so the contrast tests
whether BACE's use of correlated traits pays off. Simulation v1 (accuracy only, earlier BACE settings, equal-rates
castor) found BACE ahead at λ = 0.3 and castor ahead at λ = 1. Equal rates run to their bound at low λ and then
impute uniform classes, below the observed-frequency floor (review of 2026-09-26), which is why the flexible models
are the primary frequentist arms here.

```{r disc-data, eval = has_disc}
dd <- function(f) utils::read.csv(file.path(disc_dir, f), stringsAsFactors = FALSE)
dcl <- dd("agg_disc_cells_l.csv"); dcr <- dd("agg_disc_cells.csv"); ddn <- dd("agg_disc_down.csv"); dpr <- dd("agg_disc_paired.csv")
rp <- if (file.exists(file.path(disc_dir, "repro.csv"))) dd("repro.csv") else NULL
labd <- c(lab, freqA_er = "freq A, equal rates (v1)", freqB_er = "freq B, equal rates (v1)", mode_floor = "mode floor")
trait_ord <- c("bin", "ord", "cat3")
f3n <- function(x) f3(as.numeric(x))   # read.csv gives an all-NA column (e.g. SEs of one dataset) as logical
cell_se <- function(v, s) ifelse(is.na(v), "", ifelse(is.na(s), f3n(v), paste0(f3n(v), " (", f3n(s), ")")))
# rows x columns table; `cell` maps the aggregate rows to the printed cell; a combination without rows prints blank
wide_d <- function(df, cell, rows, arms, col = "arm", labels = labd) {
  df <- df[df[[col]] %in% arms, ]; df$cell <- cell(df)
  if ("trait" %in% rows) df$trait <- factor(df$trait, levels = trait_ord)
  w <- stats::reshape(df[, c(rows, col, "cell")], idvar = rows, timevar = col, direction = "wide")
  names(w) <- sub("^cell\\.", "", names(w)); keep <- intersect(arms, names(w))
  w <- w[do.call(order, w[rows]), c(rows, keep)]
  for (k in keep) w[[k]][is.na(w[[k]])] <- ""
  names(w)[-seq_along(rows)] <- labels[keep]; w
}
```

```{r disc-value, results = "asis", eval = has_disc}
cat("## Per-value scores\n\n")
cat("Each arm is scored from its 20 draws of every missing value: accuracy of the modal class (ties counted ",
    "fractionally), multi-class Brier score of the draw frequencies, and coverage and size of the smallest class set ",
    "holding 95% of the draws. The mode floor predicts the most frequent observed class of the dataset in every ",
    "missing cell. The tables pool ρ = 0 and 0.5 (by ρ: `data/discrete/agg_disc_cells.csv`). BACE + residual is not ",
    "shown: its extra draw changes Gaussian traits only, so its discrete draws are those of BACE as shipped.\n\n", sep = "")
arms_v <- c("freqA", "freqB", "bace", "bace_chain")
caps <- c(accuracy = "Accuracy of the modal imputed class (Monte Carlo SE), pooled over ρ. Higher is better.",
          brier = "Multi-class Brier score of the draw frequencies (Monte Carlo SE), pooled over ρ. Lower is better.",
          set_coverage = "Coverage of the smallest class set holding 95% of the 20 draws (Monte Carlo SE), pooled over ρ. Nominal 0.95.",
          set_size = "Mean number of classes in that set (Monte Carlo SE), pooled over ρ.")
for (m in names(caps)) {
  cat(tab(wide_d(dcl, function(g) cell_se(g[[m]], g[[paste0(m, "_se")]]), c("trait", "n", "lambda"),
                 if (m == "accuracy") c(arms_v, "mode_floor") else arms_v), caps[[m]]), sep = "\n")
  cat("\n\n")
}
ab <- dcr[dcr$arm == "mode_floor", ]   # one row per dataset and trait, whichever arm set holds the dataset
ab <- stats::aggregate(cbind(absent = n_fits_trait_absent, all = n_fits + n_fits_trait_absent + n_fits_missing) ~ trait + n, ab, sum)
ab <- ab[ab$absent > 0, ]
na_d <- dcr[dcr$arm != "mode_floor" & !is.na(dcr$n_na_draws) & dcr$n_na_draws > 0, ]
na_d <- if (nrow(na_d)) stats::aggregate(n_na_draws ~ arm + trait, na_d, sum) else na_d
cat(if (nrow(ab)) paste0("Traits with one class in the complete data have nothing to predict and are skipped by every arm: ",
                         paste(sprintf("`%s` in %d of %d datasets at n = %d", ab$trait, ab$absent, ab$all, ab$n), collapse = "; "),
                         ". ") else "No trait had a single class in the complete data. ",
    sprintf("Per-value rows missing for another reason (an arm or scoring error; see `data/discrete/fit_disc_cells.csv.gz`): %d. ",
            sum(dcr$n_fits_missing[dcr$arm != "mode_floor"])),
    if (nrow(na_d)) paste0("Draws returned as NA, dropped cell by cell before scoring: ",
                           paste(sprintf("%s `%s` %d", labd[na_d$arm], na_d$trait, na_d$n_na_draws), collapse = "; "), ".")
    else "No arm returned NA draws.", "\n\n", sep = "")
cat("Interpretation: to be written from the results.\n\n")
```

```{r disc-down, results = "asis", eval = has_disc}
cat("## Downstream estimate: PGLS slope of c1 on bin\n\n")
tt <- if (file.exists(file.path(disc_dir, "truth_slope_c1_bin.csv"))) dd("truth_slope_c1_bin.csv") else NULL
cat("The Rubin-pooled PGLS slope of c1 on `bin` (`bin` = yes coded 1) has two targets. The per-dataset target is ρ ",
    "times the GLS slope, at the true λ, of `bin`'s liability on `bin`: the expected complete-data estimate given the ",
    "tree and that liability. The population target is the mean complete-data estimate over ",
    if (is.null(tt)) "fresh datasets" else paste0(paste(format(sort(unique(tt$R)), big.mark = ","), collapse = " or "),
                                                  " fresh datasets (Monte Carlo SE at most ", f3n(max(tt$mc_se)), ")"),
    " per cell (`data/discrete/truth_slope_c1_bin.csv`). Neither is ρ. At λ ≤ 0.7, or ρ = 0, the two nearly agree. At ",
    "λ = 1 and ρ = 0.5 the per-dataset target varies between datasets by more than the standard error (`bin` is ",
    "almost constant within clades), so even the complete-data interval covers the population target in only 31% to ",
    "56% of datasets; coverage there is read against the per-dataset target. The complete-data column applies the ",
    "same analysis to the data before masking, with the t interval on n − 2 df.\n\n", sep = "")
dn <- ddn; dn$truth <- f3n(dn$truth); rows_d <- c("n", "lambda", "rho", "truth")
arms_d <- c("complete", "freqA", "freqB", "freqA_er", "freqB_er", "bace", "bace_chain", "bace_resid")
cat(tab(wide_d(dn, function(g) cell_se(g$coverage_cond, g$coverage_cond_se), rows_d, arms_d),
        "c1 ~ bin: coverage of the per-dataset target by the pooled 95% interval (Monte Carlo SE). Nominal 0.95."), sep = "\n"); cat("\n\n")
cat(tab(wide_d(dn, function(g) cell_se(g$bias_cond, g$bias_cond_se), rows_d, arms_d),
        "c1 ~ bin: bias of the pooled estimate against the per-dataset target (Monte Carlo SE)."), sep = "\n"); cat("\n\n")
cat(tab(wide_d(dn, function(g) cell_se(g$coverage, g$coverage_se), rows_d, arms_d),
        "c1 ~ bin: coverage of the population target (the truth column) by the pooled 95% interval (Monte Carlo SE)."), sep = "\n"); cat("\n\n")
cat(tab(wide_d(dn, function(g) ifelse(is.na(g$diff_cd_mean), "", paste0(f3n(g$diff_cd_mean), " (", f3n(g$diff_cd_sd), ")")),
               rows_d, setdiff(arms_d, "complete")),
        "c1 ~ bin: pooled estimate minus the complete-data estimate of the same dataset, mean (SD over datasets)."), sep = "\n")
cat("\n\n")
ex <- ddn[ddn$arm == "complete", ]
ex <- stats::aggregate(cbind(excluded = n_excluded, undefined = n_undefined, all = n_fits + n_excluded + n_undefined + n_missing) ~ n, ex, sum)
cat("The estimand is undefined when `bin` has one class in the complete data, or one observed class (every imputed ",
    "dataset then has a constant `bin`); those datasets are left out of every arm (",
    paste(sprintf("%d and %d of %d at n = %d", ex$excluded, ex$undefined, ex$all, ex$n), collapse = "; "), "). ",
    sprintf("Fits with no c1 ~ bin row for another reason (an arm or scoring error; see `data/discrete/fit_disc_estimands.csv.gz`): %d.\n\n",
            sum(ddn$n_missing[ddn$arm != "complete"])), sep = "")
cat("Interpretation: to be written from the results.\n\n")
```

```{r disc-paired, results = "asis", eval = has_disc}
cat("## Paired contrasts: BACE against the frequentist arms\n\n")
clab <- c("bace_chain - freqA" = "chained BACE − freq A", "bace - freqA" = "BACE as shipped − freq A",
          "bace_chain - bace" = "chained BACE − BACE as shipped", "freqA - freqB" = "freq A − freq B",
          "freqA - freqA_er" = "freq A: flexible − equal rates", "freqB - freqB_er" = "freq B: flexible − equal rates")
for (ms in c("accuracy", "brier")) {
  w <- wide_d(dpr[dpr$measure == ms, ], function(g) cell_se(g$diff, g$se), c("trait", "n", "lambda", "rho"), names(clab),
              col = "contrast", labels = clab)
  cat(tab(w, paste0("Paired per-dataset difference in ", if (ms == "accuracy") "accuracy" else "Brier score",
                    " (SE over datasets; counts in `data/discrete/agg_disc_paired.csv`). ",
                    if (ms == "accuracy") "Positive favours the first arm." else "Negative favours the first arm.")),
      sep = "\n"); cat("\n\n")
}
cat("Interpretation: to be written from the results.\n\n")
```

```{r disc-repro, results = "asis", eval = has_disc}
cat("## Reproduction of the continuous results\n\n")
# an exact match is possible only where the same cluster (same R and BACE builds) ran both the stored and the new
# fit; a pair of clusters gives an independent replicate
rb <- if (!is.null(rp)) rp[rp$set == "bace" & rp$arm == "all" & rp$host != "all", ] else NULL
if (is.null(rb) || !nrow(rb)) {
  cat("No BACE fit of the discrete re-run was found in the continuous pool, so the stored continuous rows were not re-checked.\n\n")
} else {
  same <- rb[rb$host == rb$cont_host, ]; cross <- rb[rb$host != rb$cont_host, ]
  cat(sprintf(paste("BACE fits re-run on the cluster that ran the stored fit: %d, of which %d reproduced every stored",
                    "continuous row (pooled estimates and SEs, per-value z-RMSE and coverage) to within %g. Fits re-run on",
                    "another cluster: %d; they are independent replicates of the stored fits (largest difference in a",
                    "pooled estimate %s). Details by cluster pair and arm: `data/discrete/repro.csv`.\n\n"),
              sum(same$n_fits), sum(same$n_fits_identical), rb$tol[1], sum(cross$n_fits),
              if (nrow(cross)) formatC(max(cross$max_abs_estimate, na.rm = TRUE), format = "g", digits = 2) else "none"))
}
```

# Limitations

- One mechanism: Brownian motion with Pagel's λ, MCAR 30%, random coalescent trees. No MAR or clade-structured
  missingness, no OU, no real trees.
- The frequentist model shares one λ across its traits; prp carries extra non-phylogenetic noise, so its fitted
  variance is inflated at λ near 1 (c1 and c2 are unaffected).
- BACE at n = 1000 has 100 datasets per cell, so its Monte Carlo SEs are about 1.4 times the others'.
- At λ = 1 and n = 1000 freq A and chained BACE both cover slightly above 0.95; not investigated.

# Files and reproduction

| file | contents |
|---|---|
| `data/fits.csv` | one row per fit: cell, seed, cluster, node, time, wall times, errors, versions |
| `data/fit_cells.csv.gz` | fit × arm × trait: per-value coverage, width, z-RMSE, interval score |
| `data/fit_estimands.csv.gz` | fit × arm × estimand: pooled estimate, SE, interval, df, FMI, truth, covered |
| `data/bace_diagnostics.csv` | BACE convergence verdict, drift, ESS, input cleaning |
| `data/failure_records.csv` | first-pass BACE failures kept as records |
| `data/agg/` | aggregated tables used above, and `report.json` for the web pages |
| `data/discrete/` | discrete results (to come) |

Pipeline (worktree root): `script/rubin_cell.R` (one dataset, all arms of a set) run on the clusters with
`script/rubin_campaign_nibi.sh` and `script/rubin_totoro.sh`; `script/rubin_pool.sh` copies results to the pool;
`script/rubin_missing.R` checks completeness; `script/rubin_campaign_aggregate.R` aggregates;
`script/rubin_study/build_data.R` writes this folder's tables; `quarto render script/rubin_study/study.qmd`.
Tests: `script/tests-rubin/`. The raw per-fit files (556 MB) stay in the pool and on the clusters.
