---
title: "Frequentist and Bayesian multiple imputation of phylogenetic traits under Rubin's rules: a simulation study"
subtitle: "Methods and results for continuous and discrete traits"
date: last-modified
format:
  html:
    minimal: true
    css: study.css
    toc: true
    toc-depth: 3
    embed-resources: true
    number-sections: true
    fig-cap-location: bottom
    tbl-cap-location: top
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
d <- "data"; ag <- file.path(d, "agg_pig"); disc_dir <- file.path(d, "discrete")
rd <- function(f) utils::read.csv(file.path(ag, f), stringsAsFactors = FALSE)
dd <- function(f) utils::read.csv(file.path(disc_dir, f), stringsAsFactors = FALSE)
cells <- rd("cells.csv"); cells_l <- rd("cells_l.csv"); down <- rd("down.csv"); down_l <- rd("down_l.csv"); down_n <- rd("down_n.csv")
failures <- rd("failures.csv")
fits <- utils::read.csv(file.path(d, "fits.csv"), stringsAsFactors = FALSE)
diag <- utils::read.csv(file.path(d, "bace_diagnostics.csv"), stringsAsFactors = FALSE)
dfits <- dd("fits.csv"); dcl <- dd("agg_disc_cells_l.csv"); ddn <- dd("agg_disc_down.csv")
dpr <- dd("agg_disc_paired.csv"); rp <- dd("repro.csv"); tt <- dd("truth_slope_c1_bin.csv")
# display labels: the frequentist methods are Rphylopars (continuous) and castor (discrete)
lab <- c(complete = "complete data", freqA = "frequentist, proper", freqB = "frequentist, fixed parameters",
         bace = "BACE as shipped", bace_chain = "chained BACE", bace_resid = "BACE + extra residual (control)",
         pig_post = "pigauto posterior")
labd <- c(lab, freqA_er = "frequentist proper, equal rates", mode_floor = "most frequent class")
ord <- names(lab)
f3 <- function(x) formatC(as.numeric(x), format = "f", digits = 3)
f2 <- function(x) formatC(as.numeric(x), format = "f", digits = 2)
big <- function(x) format(x, big.mark = ",")
rng <- function(x, f = f3) { x <- x[is.finite(x)]; a <- f(min(x)); b <- f(max(x)); if (a == b) a else paste(a, "to", b) }
# coverage MCSE for one interval per dataset: binomial, floored at the nominal rate so a coverage of 1.00 keeps an SE
cse <- function(p, r) sqrt(pmax(p * (1 - p), 0.95 * 0.05) / r)
wide <- function(df, value, se = NULL, rows = c("n", "lambda"), arms = ord) {
  df <- df[df$arm %in% arms, ]
  df$cell <- if (is.null(se)) f3(df[[value]]) else paste0(f3(df[[value]]), " (", f3(df[[se]]), ")")
  w <- stats::reshape(df[, c(rows, "arm", "cell")], idvar = rows, timevar = "arm", direction = "wide")
  names(w) <- sub("^cell\\.", "", names(w))
  keep <- intersect(arms, names(w)); w <- w[do.call(order, w[rows]), c(rows, keep)]; names(w)[-seq_along(rows)] <- lab[keep]
  names(w)[names(w) == "lambda"] <- "λ"; names(w)[names(w) == "rho"] <- "ρ"; w
}
library(ggplot2)
knitr::opts_chunk$set(dev = "ragg_png", dpi = 200)
pal <- c("complete data" = "#7b8580", "frequentist, proper" = "#d4891a", "frequentist, fixed parameters" = "#b565a0",
         "BACE as shipped" = "#c4452b", "chained BACE" = "#2c6b59", "BACE + extra residual (control)" = "#5b82c0",
         "pigauto posterior" = "#3b3b98")
theme_rubin <- theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        strip.text = element_text(face = "bold", hjust = 0, size = 12), legend.position = "bottom",
        legend.text = element_text(size = 10), plot.background = element_rect(fill = "white", colour = NA),
        panel.spacing = unit(1.2, "lines"))
# coverage by n, faceted by lambda; shape distinguishes rho; whiskers +/- 2 Monte Carlo SE
figcov <- function(s, y, se, ylim, shape = TRUE) {
  s$y <- y; s$lo <- pmax(y - 2 * se, 0); s$hi <- pmin(y + 2 * se, 1)
  s$method <- factor(lab[s$arm], levels = lab); s$lam <- factor(paste("λ =", s$lambda))
  s$rho_f <- factor(paste("ρ =", s$rho))
  g <- ggplot(s, aes(factor(n), y, colour = method, group = interaction(method, rho_f))) +
    geom_hline(yintercept = 0.95, linetype = "dashed", colour = "#5a625e") +
    geom_pointrange(if (shape) aes(ymin = lo, ymax = hi, shape = rho_f) else aes(ymin = lo, ymax = hi),
                    position = position_dodge(width = 0.75), size = 0.3, linewidth = 0.5) +
    facet_wrap(~lam) + scale_colour_manual(values = pal, name = NULL) +
    coord_cartesian(ylim = ylim) + labs(x = "number of species n", y = "coverage", shape = NULL) + theme_rubin +
    guides(colour = guide_legend(nrow = 2, order = 1), shape = guide_legend(order = 2))
  if (shape) g + scale_shape_manual(values = c(16, 17)) else g
}
TRN <- c(bin = "binary", ord = "ordinal", cat3 = "three-class")
trait_ord <- names(TRN)
cell_se <- function(v, s) ifelse(is.na(v), "", ifelse(is.na(s), f3(v), paste0(f3(v), " (", f3(s), ")")))
wide_d <- function(df, cell, rows, arms, col = "arm", labels = labd) {
  df <- df[df[[col]] %in% arms, ]; df$cell <- cell(df)
  if ("trait" %in% rows) df$trait <- factor(TRN[df$trait], levels = TRN)
  w <- stats::reshape(df[, c(rows, col, "cell")], idvar = rows, timevar = col, direction = "wide")
  names(w) <- sub("^cell\\.", "", names(w)); keep <- intersect(arms, names(w))
  w <- w[do.call(order, w[rows]), c(rows, keep)]
  for (k in keep) w[[k]][is.na(w[[k]])] <- ""
  names(w)[-seq_along(rows)] <- labels[keep]; names(w)[names(w) == "lambda"] <- "λ"; names(w)[names(w) == "rho"] <- "ρ"; w
}
# ---- key numbers quoted in the text (computed here, never typed)
dn1 <- function(a, n, e, v) down_n[down_n$arm == a & down_n$n == n & down_n$estimand == e, v]
dl1 <- function(a, n, l, e, v) down_l[down_l$arm == a & down_l$n == n & down_l$lambda == l & down_l$estimand == e, v]
lowcov <- cells_l$coverage[cells_l$arm == "bace" & cells_l$lambda < 1]
# paired per-value accuracy: chained BACE minus freq A, per dataset (mean of c1 and c2), by n and lambda
fcell <- utils::read.csv(gzfile(file.path(d, "fit_cells.csv.gz")), stringsAsFactors = FALSE)
fcell <- fcell[fcell$trait %in% c("c1", "c2") & fcell$arm %in% c("freqA", "bace_chain"), ]
zz <- stats::aggregate(zRMSE ~ tag + n + lambda + rho + arm, fcell, mean)
zw <- stats::reshape(zz, idvar = c("tag", "n", "lambda", "rho"), timevar = "arm", direction = "wide")
zw$d <- zw$zRMSE.bace_chain - zw$zRMSE.freqA; zw <- zw[is.finite(zw$d), ]
zp <- stats::aggregate(d ~ n + lambda, zw, function(x) c(mean = mean(x), se = sd(x) / sqrt(length(x)), k = length(x)))
zp <- data.frame(n = zp$n, lambda = zp$lambda, diff = zp$d[, "mean"], se = zp$d[, "se"], k = zp$d[, "k"])
pa <- dpr[dpr$contrast == "bace_chain - freqA" & dpr$measure == "accuracy", ]
lead <- pa$diff[pa$lambda < 1]; leadz <- (pa$diff / pa$se)[pa$lambda < 1]
ddv <- function(a, l, r, n, v) ddn[ddn$arm == a & ddn$lambda %in% l & ddn$rho %in% r & ddn$n %in% n, v]
fc <- utils::read.csv(gzfile(file.path(disc_dir, "fit_disc_cells.csv.gz")), stringsAsFactors = FALSE)
bfail <- unique(fc$tag[fc$arm == "bace" & fc$status == "missing" & grepl("bace_fit", fc$errors)])
cfail <- unique(fits$tag[fits$set == "bace" & grepl("bace_fit", fits$errors)])
cfail300 <- unique(fits$tag[fits$set == "bace" & grepl("bace_fit", fits$errors) & fits$n <= 300])
chain_only_c <- sum(fits$set == "bace" & grepl("bace_chain", fits$errors) & !grepl("bace_fit", fits$errors))
chain_only_d <- length(unique(fc$tag[fc$arm == "bace_chain" & fc$status == "missing" & !grepl("bace_fit", fc$errors) & !is.na(fc$errors) & nzchar(fc$errors)]))
rb <- rp[rp$set == "bace" & rp$arm == "all" & rp$host != "all", ]
same <- rb[rb$host == rb$cont_host, ]; cross <- rb[rb$host != rb$cont_host, ]
l1clean <- tapply(!is.na(diag$input_fix[diag$lambda == 1]) & nzchar(diag$input_fix[diag$lambda == 1]), diag$n[diag$lambda == 1], mean)
h1000 <- fits$wall_s_total[fits$set == "bace" & fits$n == 1000 & !grepl("bace_fit", fits$errors)] / 3600
dv <- function(a, n, l, r, e, v = "coverage") down[down$arm == a & down$n %in% n & down$lambda %in% l & down$rho %in% r & down$estimand == e, v]
# median wall time of one step of a fit (wall_s holds "step=seconds;..."), in hours
wpart <- function(set, key, n) { s <- fits$wall_s[fits$set == set & fits$n == n & grepl(paste0(key, "="), fits$wall_s)]
  stats::median(as.numeric(sub(paste0(".*", key, "=([0-9.]+).*"), "\\1", s))) / 3600 }
s5 <- down[down$estimand == "slope" & down$rho == 0.5, ]
cshift <- merge(s5[s5$arm == "bace_chain", c("n", "lambda", "bias")], s5[s5$arm == "bace", c("n", "lambda", "bias")], by = c("n", "lambda"))
cshift <- cshift$bias.x - cshift$bias.y
# pigauto posterior MI (pig_post): paired differences against proper frequentist MI, and the fit record
pr <- rd("paired.csv")
pp <- function(e, n, l, v = "diff") pr[pr$contrast == "pig_post - freqA" & pr$measure %in% e & pr$n %in% n & pr$lambda %in% l, v]
cells_n <- rd("cells_n.csv")
cn1 <- function(a, n, v = "coverage") cells_n[cells_n$arm == a & cells_n$n == n, v]
pf <- failures[failures$set == "pig", ]
```

# Abstract {.unnumbered}

Multiple imputation (MI) is the standard way to analyse comparative data with missing trait values. Its intervals
are valid only if the imputations are *proper*: independent draws of the missing values that also carry
uncertainty about the imputation model's parameters. We tested BACE, a Bayesian chained-equation imputer built on
MCMCglmm, against proper and fixed-parameter frequentist imputation in a factorial simulation of phylogenetic trait
data (18 scenarios: n = 100, 300, 1000 species; Pagel's λ = 0.3, 0.7, 1; trait correlation ρ = 0, 0.5; 200
datasets per scenario; M = 20 imputations), scoring every method with Rubin's rules. For continuous traits
(`r big(sum(fits$set == "bace"))` BACE and `r big(sum(fits$set == "freq"))` frequentist fits), BACE as shipped was
not proper. Its 20 final imputations all start from one dataset whose missing entries hold single posterior-mean
fills, so they share those fills and are too alike: interval coverage fell to `r rng(lowcov, f2)` for single values
at λ ≤ 0.7 and to `r f2(dv("bace", 1000, 0.7, 0, "cor"))` for the phylogenetic correlation (n = 1000,
λ = 0.7, ρ = 0), against a nominal 0.95. Starting each final imputation from the previous one ("chaining") restored
coverage to nominal or above (conservative at n = 1000) in every scenario except complete phylogenetic signal with
correlated traits at n = 1000, where chained BACE and proper frequentist MI both covered the correlation
`r f3(dv("bace_chain", 1000, 1, 0.5, "cor"))`; chained BACE was then more accurate per value than proper frequentist MI at n ≥ 300 and λ ≤ 0.7, and less accurate at n = 100. pigauto's posterior MI, a
multivariate phylogenetic mixed model sampled by Markov chain Monte Carlo (`r big(sum(pf$files))` fits on the same
continuous datasets), covered like proper frequentist MI at n ≥ 300 and λ ≤ 0.7 and slightly less at n = 100; with
complete phylogenetic signal and correlated traits (λ = 1, ρ = 0.5) it pulled the slope toward zero at every n, and
at n = 1000 its slope interval covered `r f2(dv("pig_post", 1000, 1, 0.5, "slope"))` against
`r f2(dv("freqA", 1000, 1, 0.5, "slope"))`. For discrete traits (binary, ordinal, three-class; n = 100 and 300; the same
datasets), chained BACE's imputed classes were `r rng(lead, f2)` more accurate than those of castor, the frequentist
tool that imputes one trait at a time, at λ ≤ 0.7. Imputing a trait without the other variable of a later
regression biased that regression: at ρ = 0.5 and λ ≤ 0.7 the slope of a continuous trait on a binary trait was
pulled toward zero and its coverage fell to `r rng(ddv("freqA", c(0.3, 0.7), 0.5, 300, "coverage_cond"), f2)` at
n = 300. We recommend chaining BACE's final imputations and imputing every variable of the planned analysis jointly.

# Introduction

Trait databases used in comparative biology are incomplete, and analysing only complete cases discards data and
can bias estimates (Nakagawa & Freckleton 2008). MI replaces each missing value with M plausible values drawn from
its predictive distribution, analyses each completed dataset, and combines the M results with Rubin's rules into
one estimate and one interval (Rubin 1987). The combined interval has its nominal coverage (the share of 95%
intervals that contain the true value should be 0.95) when the imputations are proper: over repeated datasets, the
mean of the M estimates is unbiased for the complete-data estimate and their spread is unbiased for its extra
variance. Independent draws from the predictive distribution under a correct model, with the model's parameters
drawn as well, satisfy this.

Phylogenetic imputation adds a dependency: species share ancestry, so a missing value is predicted from relatives
as well as from other traits. Frequentist tools fit a phylogenetic model once and draw from it: Rphylopars for
continuous traits (Goolsby, Bruggeman & Ané 2017) and castor for discrete traits under an Mk model (Louca &
Doebeli 2018). BACE imputes continuous and discrete traits together by chained equations: it cycles through the
traits, modelling each on all the others with a phylogenetic random effect in MCMCglmm (Hadfield 2010).

We asked two questions. First, are BACE's imputations proper, judged by Rubin-pooled coverage against a proper and
a fixed-parameter frequentist method on continuous data close to the frequentist model? Second, when traits are
discrete, does BACE's use of the other traits improve imputation relative to the frequentist tool a user can
actually run? The design follows the ADEMP framework (aims, data-generating mechanisms, estimands, methods,
performance measures; Morris, White & Crowther 2019) and the reporting items of Williams et al. (2024).

::: {.box .box-note}
[What to do, for users of MI with phylogenetic data]{.box-title}

1. If you use BACE, chain its final imputations: run its final step once per imputation, each time starting from
   the previous imputation (@sec-methods; code in `script/rubin_bace.R`, `mi_bace_chain()`). On one core, BACE's
   fit took a median of `r f2(wpart("bace", "bace_fit", 100))`, `r f2(wpart("bace", "bace_fit", 300))` and
   `r f2(wpart("bace", "bace_fit", 1000))` hours at n = 100, 300 and 1000, and chaining added
   `r f2(wpart("bace", "bace_chain", 100))`, `r f2(wpart("bace", "bace_chain", 300))` and
   `r f2(wpart("bace", "bace_chain", 1000))` hours.
2. For continuous traits only, proper frequentist MI (Rphylopars, refitted before each draw) covered at or near
   nominal except with complete phylogenetic signal (λ = 1) and correlated traits, where its shared λ biased the
   estimates, and took a median of `r f2(60 * wpart("freq", "freqA", 100))`, `r f2(60 * wpart("freq", "freqA", 300))`
   and `r f2(60 * wpart("freq", "freqA", 1000))` minutes per dataset.
3. pigauto's posterior MI (`draws_method = "posterior"`) covered like proper frequentist MI except with complete
   phylogenetic signal and correlated traits, where it pulled the slope toward zero and its intervals under-covered
   at n ≥ 300 (@sec-pig).
4. Impute every variable of your planned analysis in one model. Imputing a trait without the other variable of a
   later regression biases the slope toward zero, and more data makes the interval more confidently wrong.
5. Build intervals with Rubin's rules, never from the 2.5th and 97.5th percentiles of your imputations: with
   M = 20 that interval covers only about 87% of the time.
:::

::: {.box .box-terms}
[Terms used in this report]{.box-title}

- **Coverage**: the share of 95% intervals, over repeated simulated datasets, that contain the true value. It should
  be 0.95; below means intervals are too narrow (overconfident), above means too wide.
- **Proper imputation**: imputations that vary as much as the missing values are truly uncertain, including
  uncertainty about the imputation model's own parameters. Only then do Rubin's rules give valid intervals.
- **Rubin's rules**: average the estimates from the M completed datasets, and add their spread to the average
  within-dataset variance to get the total uncertainty.
- **Monte Carlo SE**: the uncertainty of a simulation result that comes from running a finite number of datasets
  (here 200); a difference of less than about two Monte Carlo SEs could be chance.
- **Liability**: an unobserved continuous value behind a discrete trait; the trait is "yes" when the liability
  passes a threshold.
- **Mk model**: a model in which a discrete trait jumps between its states along the branches of the tree at
  estimated rates. Its **stationary distribution** is the mix of states it settles into after a long time; when the
  rates are very fast the tree carries no information and draws come from that mix.
- **Parametric bootstrap**: simulate new data from the fitted model and refit it, to see how much the parameter
  estimates could vary.
- **REML, Fisher's z, whitening**: standard devices. REML estimates variance parameters with less small-sample
  bias; Fisher's z is a transformation that makes a correlation's sampling distribution close to normal; whitening
  removes the phylogenetic dependence among species before a correlation is computed.
- **Forward filtering, backward sampling**: an exact way to draw the states of all unobserved species at once, so
  that the draws respect the tree and each other.
- **Per-dataset target**: the slope the complete data would give on average for this particular tree and these
  particular liabilities; the value each method's estimate is compared with for the slope of c1 on bin.
- **Most frequent class (mode floor)**: the accuracy of always guessing the commonest observed class; an imputer
  that does worse than this learned nothing useful.
- **Uncongeniality**: the imputation model and the analysis model disagree about which variables are related; Rubin's
  intervals can then be too wide or too narrow.
:::

# Methods {#sec-methods}

## Aims

1. **Primary.** Estimate the coverage and bias of Rubin-pooled estimates for BACE as shipped and for proper and
   fixed-parameter frequentist MI on continuous traits.
2. **Secondary.** Compare the accuracy and calibration of imputed classes, and the Rubin-pooled estimate of an
   association involving a discrete trait, between BACE and castor on the same datasets.

3. **Added later.** Estimate the same coverage and bias for pigauto's posterior MI on the continuous traits.

The chained variant of BACE and the extra-residual control were added after a pre-run showed BACE as shipped
under-covering, and they were evaluated on the same seeds; they were not validated on fresh seeds (Limitations).
pigauto's posterior MI was run after the BACE and frequentist campaign, on the same datasets and seeds.

## Data-generating mechanism

Each dataset is built from n, λ, ρ and an integer seed. For a given n and seed, the tree and the underlying random
numbers are the same in all six (λ, ρ) scenarios, so comparisons between scenarios use common random numbers.

- **Tree.** `ape::rcoal(n)` (Paradis & Schliep 2019), rescaled to a root-to-tip depth of 1; *C* is its tip
  correlation matrix.
- **Liabilities.** Each trait comes from an unobserved continuous value, its *liability*, that evolves along the
  tree by Brownian motion. Formally, eight liabilities *L* = *V*~λ~^1/2^ *Z* Σ^1/2^, with *Z* an n × 8 matrix of
  independent standard normals, *V*~λ~ = λ*C* + (1 − λ)*I* (Pagel's λ; Pagel 1999), and Σ a correlation matrix
  with every off-diagonal equal to ρ among the seven trait columns and 0.35 between the eighth (driver) column and
  each trait. Every liability has unit variance and phylogenetic signal λ.
- **Traits.** c1 = *L*~1~ and c2 = *L*~2~ (continuous); cnt ~ Poisson(exp(1.5 + 0.8*L*~3~)) (count);
  prp = logistic(*L*~4~ + ε), ε ~ N(0, 0.3²), truncated to [10^−4^, 1 − 10^−4^] (proportion);
  bin = 1(*L*~5~ > 0) (binary); ord cuts *L*~6~ at the standard-normal quartiles (four ordered levels);
  cat3 cuts *L*~7~ at the standard-normal terciles (three classes, ordered by construction); d1 = *L*~8~, a trait
  that is always measured, like body mass in a real database. Thresholds are fixed at population quantiles.
- **Missingness.** Missing completely at random (MCAR): for each of the seven traits, ⌈0.3 n⌉ entries (30, 90
  and 300) are masked at random; d1 is never missing.
- **Scenarios.** Full factorial over n ∈ {100, 300, 1000}, λ ∈ {0.3, 0.7, 1} and ρ ∈ {0, 0.5}. The levels span weak
  to complete phylogenetic signal, no and moderate trait correlation, and 30% missingness per trait.
- **Replicates.** Seeds 1 to 200 in every scenario. With 200 datasets the Monte Carlo SE of a downstream coverage
  of 0.95 is 0.015, so a departure of 0.03 is detectable at two SEs; per-value coverage, averaged over 60 to 600
  entries per dataset, is far more precise. BACE at n = 1000 used 100 datasets per scenario (SE 0.022), because
  a fit took a median of `r f2(median(h1000))` hours (90% between `r f2(quantile(h1000, 0.05))` and
  `r f2(quantile(h1000, 0.95))`).
- **Pairing.** Each run rebuilds its dataset from (n, λ, ρ, seed) with R's L'Ecuyer-CMRG generator, so every
  method sees the same datasets; complete-data estimates agreed exactly across all result files. Each method
  re-seeds from the dataset seed plus a fixed offset, so its draws do not depend on which methods ran first.

Under this mechanism c2 = ρ c1 + *e* with *e* ~ N(0, (1 − ρ²)*V*~λ~). A phylogenetic generalised least squares
(PGLS) regression of c2 on c1 with Pagel's λ is correctly specified, and both the population slope and the
phylogenetic correlation of c1 and c2 equal ρ.

## Estimands and targets

1. **Each missing value.** The true value of each masked c1 and c2 (prp on the logit scale, not reported here),
   and the true class of each masked bin, ord and cat3.
2. **PGLS slope of c2 on c1**, Pagel's λ estimated by REML on [0, 1]. Target: ρ.
3. **Phylogenetic correlation of c1 and c2**, both traits whitened with the λ of the slope fit, on Fisher's z with
   variance 1/(n − 3). Target: ρ.
4. **PGLS slope of c1 on bin** (bin = 1 for "yes"). Regressing a continuous trait on a 0/1 trait does not give ρ:
   the expected slope depends on the tree. The **per-dataset target** is ρ times the GLS slope, at the true λ, of
   *L*~5~ on bin: the expected complete-data estimate given the tree and the binary trait's liability, when λ is
   known. Because it conditions on *L*~5~, which carries more information than bin, the complete-data interval
   is expected to cover it at the nominal rate at ρ = 0 and slightly conservatively at ρ = 0.5 (observed
   `r rng(ddv("complete", c(0.3, 0.7, 1), 0, c(100, 300), "coverage_cond"))` and
   `r rng(ddv("complete", c(0.3, 0.7, 1), 0.5, c(100, 300), "coverage_cond"))`), and methods are compared with
   the complete-data row rather than with 0.95. A **population target**, the mean complete-data estimate over
   `r big(unique(tt$R))` fresh datasets per scenario (Monte Carlo SE at most `r f3(max(tt$mc_se))`), agrees
   closely except at λ = 1 and ρ = 0.5, where the tree-specific slope varies between datasets by more than its
   standard error and even complete data cover the population target only 30% to 56% of the time. Coverage is
   therefore read against the per-dataset target; population-target coverage is in `data/discrete/agg_disc_down.csv`.

## Methods compared

Every method produced M = 20 completed datasets for every dataset.

- **Frequentist, proper (continuous: Rphylopars).** One joint fit of c1, c2, logit(prp) and d1 with
  `Rphylopars::phylopars` (Pagel's λ shared by the four traits, correlated phylogenetic covariance, REML). For
  each imputation: simulate a complete dataset from the fitted multivariate normal, mask it like the real data,
  refit, and draw the missing entries jointly from their conditional normal distribution given the observed
  entries under the refitted parameters. This parametric bootstrap stands in for the posterior of the parameters,
  so the scheme is approximately proper. A refit is rejected and redrawn (at most five times) if it fails, if λ
  leaves [0, 1] or the covariance is not positive definite, if its implied tip variance lies outside 1/50 to 50
  times the original fit's, or if its conditional distribution is degenerate; the original fit must itself imply
  a tip variance within 10^4^ times the observed variance. Rejections were rare (the failures table). The count
  and discrete traits are not in this model.
- **Frequentist, fixed parameters (continuous: Rphylopars).** The same fit used once: 20 conditional draws with
  the parameters held at their estimates. This is improper and serves as a contrast.
- **BACE as shipped.** BACE 0.1.0 used as installed: one formula per trait on all other traits with a phylogenetic
  random effect, 5 runs of 50,000 iterations (burn-in 10,000, thinning 25), one-vs-rest models for cat3, and
  n~final~ = 20 (launch: `script/rubin_campaign_nibi.sh`, `RUNS=5 NITT=50000`). BACE first cycles through the
  traits until the imputations settle, filling missing entries with posterior means. It then makes its 20 final
  imputations; each draws one posterior iteration plus a residual for every trait, but every one starts from the
  same settled dataset, so when a trait is imputed, the other traits' missing entries are held at the same fills
  in all 20. Before fitting, unused factor levels are dropped and a discrete trait with fewer than two observed
  classes is left out; otherwise BACE stops at λ = 1. At λ = 1 this cleaning was needed in about half of the
  fits (`r paste(round(100 * l1clean), collapse = "%, ")`% at n = 100, 300, 1000).
- **Chained BACE.** BACE's own final step (`bace_final_imp`, n~final~ = 1) run 20 times, each starting from the
  previous imputation instead of the shared dataset; same fit and settings, so it is paired with BACE as shipped.
  Successive chained imputations are consecutive sweeps of one Markov chain, not independent draws, and the first
  starts from the shared dataset.
- **BACE + extra residual (control).** BACE as shipped plus a second, independent residual draw on the Gaussian
  traits. It adds noise without removing the shared start, to show that extra noise alone is not the cure.
- **pigauto, posterior (continuous).** `pigauto::multi_impute(draws_method = "posterior", m = 20)` (pigauto commit
  b565cad) on the same block as the proper frequentist method: c1, c2, logit(prp) and d1, on the same scale
  (`log_transform = FALSE`). The block follows a multivariate phylogenetic mixed model,
  vec(*Y*) ~ N(**1**μ^⊤^, Σ~P~ ⊗ *R* + Σ~E~ ⊗ *I*~n~), with *R* the tree's correlation matrix and full 4 × 4
  phylogenetic (Σ~P~) and residual (Σ~E~) covariance matrices, so each trait has its own implied λ,
  Σ~P,kk~ / (Σ~P,kk~ + Σ~E,kk~). Priors: a parameter-expanded inverse-Wishart on Σ~P~ and
  Σ~E~ ~ IW(K + 1, 0.01 diag(*s*^2^)), with *s*^2^ the observed variances. A Markov chain Monte Carlo sampler draws
  the parameters and the missing entries jointly; the 20 datasets are posterior predictive draws spaced across the
  kept sweeps, so the scheme is proper. Chains are extended when any split R-hat is at least 1.05 or any bulk
  effective sample size at most 400, up to 6 times at n = 100 and 3 times otherwise; a fit that still fails this
  rule is kept, scored and flagged as unconverged. Only the block is imputed, so this method enters the continuous
  comparisons only. Code: `script/rubin_pigauto.R`.
- **Frequentist, discrete (castor).** For each discrete trait separately, `castor::hsp_mk_model` fits an Mk model,
  the standard model for a discrete trait that jumps between states along branches at estimated rates. We used
  flexible rates (all rates different for bin and cat3; between adjacent levels only, each rate different, for
  ord), the observed class frequencies as root prior, and three optimiser starts. Missing species are filled by
  drawing their states jointly from the tree (forward filtering, backward sampling, with transition probabilities
  from the matrix exponential). The fixed-parameter version draws 20 times at the fitted rates. The proper version,
  for each imputation, simulates the trait on the tree at the fitted rates, keeps the observed species, refits,
  and draws once at the refitted rates; a refit under which the observed data are impossible is rejected and
  redrawn, bootstrap samples are redrawn until at least two classes are observed, and the root prior is not
  bootstrapped, all of which slightly understate parameter uncertainty. castor draws fill the frequentist datasets
  after their continuous draws, under separate seeds. A version with equal transition rates, as in our pilot, is
  reported for comparison. A discrete trait with one observed class is imputed as that class by every method.
- **Complete data.** The same analyses on the data before masking: the reference each method should reach.

No off-the-shelf phylogenetic frequentist method imputes discrete traits jointly with continuous ones, so the
discrete comparison is between BACE and the trait-by-trait tool available to a user. The frequentist datasets
therefore impute bin without c1 (castor) and c1 without bin (Rphylopars).

## Pooling (Rubin's rules)

For a downstream estimate with per-imputation estimates *Q̂*~m~ and variances *U*~m~, the pooled estimate is
*Q̄* = mean(*Q̂*~m~), with within-imputation variance *W* = mean(*U*~m~), between-imputation variance
*B* = var(*Q̂*~m~) and total variance *T* = *W* + (1 + 1/M)*B*. Slope intervals use Barnard and Rubin's (1999)
degrees of freedom with complete-data degrees of freedom n − 2. The correlation is pooled on Fisher's z with
complete-data degrees of freedom treated as infinite (Rubin's 1987 degrees of freedom) and back-transformed. The
fraction of missing information (roughly, the share of the uncertainty due to the missing data) is reported. For a
single missing continuous value *W* = 0, so its interval is the mean of the 20 draws ± *t*~19~ √((1 + 1/M)*B*); if
the truth and the draws are exchangeable normal variables this is an exact 95% prediction interval, so per-value
coverage tests whether each value's predictive distribution is calibrated, not whether the imputations are
proper. A percentile interval of 20 draws (R's default quantile type 7) covers only 0.872. PGLS slopes are fitted
by an exact eigenbasis REML estimator, equal to `nlme::gls` with `corPagel` to within 10^−4^.

## Performance measures

- **Continuous, per value.** Coverage of the Rubin interval; mean width and z-RMSE (RMSE of the pooled mean), both
  in units of the trait's SD over observed entries; interval score, width + (2/α)(lower − *y*)1(*y* < lower)
  + (2/α)(*y* − upper)1(*y* > upper) with α = 0.05 (Gneiting & Raftery 2007). Averaged within a dataset over c1 and
  c2, then over datasets.
- **Discrete, per value**, from the 20 draws: accuracy of the most frequent imputed class (ties counted
  fractionally); the multi-class Brier score Σ~k~(*p̂*~k~ − 1{*y* = *k*})² of the draw frequencies (Brier 1950),
  which carries the same finite-M term of about (1 − Σ~k~ *p*~k~²)/20 for every method; coverage and size of the
  smallest set of classes holding 95% of the draws (ties at its edge counted fractionally), which for a calibrated
  imputer with 20 draws is at least 0.95 (0.96 to 1.00 for a binary trait), so it is compared between methods
  rather than against 0.95; and the accuracy of always guessing the most frequent observed class.
- **Downstream.** Coverage of the target, bias, RMSE, mean interval width, fraction of missing information, and
  the paired difference between each method's pooled estimate and the complete-data estimate.
- **Monte Carlo error.** Aggregates are means over datasets. For per-value scores the SE is SD/√(datasets) of the
  dataset means (entries within a dataset share one fit). For downstream coverage (one interval per dataset) the
  SE is √(*p*(1 − *p*)/datasets), with *p* floored at 0.95 so that a coverage of 1.00 keeps an SE. Methods are
  compared on the same datasets by paired per-dataset differences.
- **Failures.** A fit that stops is recorded and not scored; failures are reported by scenario. A trait with one
  class in the complete data has nothing to predict and is left out of the discrete scores for every method.

## Computing and reproducibility

Every fit ran on one core (`OPENBLAS_NUM_THREADS = 1`) on the Digital Research Alliance of Canada clusters fir,
nibi and rorqual (Slurm job arrays) and on a University of Alberta server (at most 150 cores). Single-core time was
about `r big(round(sum(fits$wall_s_total[fits$set == "bace"]) / 3600))` hours for BACE and
`r big(round(sum(fits$wall_s_total[fits$set == "freq"]) / 3600))` for the frequentist methods in the continuous
part, and `r big(round(sum(dfits$wall_s_total[dfits$set == "bace"]) / 3600))` and
`r big(round(sum(dfits$wall_s_total[dfits$set == "freq"]) / 3600))` in the discrete part (supplementary table).
The `r big(sum(pf$files))` pigauto fits ran on the University of Alberta server, 110 at a time, in 11 h 21 min of
wall time; the complete-data estimates in their result files equal those of the frequentist files for all 7,200
pairs of the same dataset and estimand. Peak memory of one BACE fit was 2.4 GB at n = 100, 6 GB at n = 300 and 19 GB at n = 1000. Software: R 4.5.0 on the
clusters and 4.5.3 on the server, BACE 0.1.0 (labelled 0.0.0.9000 on the server, with its fitting functions
checked byte-identical), MCMCglmm 2.36, Rphylopars 0.3.10, castor 1.8.7, ape 5.8.1 (versions from the stored
session information). Every result file stores its session information, the true data and the mask; the discrete
runs also store all 20 imputations of every method, so every score can be recomputed without refitting.

The discrete part re-ran BACE. Re-runs on the cluster that produced the stored continuous fit reproduced
`r sum(same$n_fits_identical)` of `r sum(same$n_fits)` fits exactly; re-runs on another cluster differed (median
absolute difference in a pooled estimate `r rng(cross$median_abs_estimate)`, maximum
`r f2(max(cross$max_abs_estimate))`) and are independent replicates. The frequentist discrete datasets were also
re-run; their continuous draws matched the continuous results of record exactly in 2,236 of 2,400 datasets and
differed in the rest through machine-level numerical differences.

# Results

## Continuous traits

### Single missing values

BACE as shipped under-covered single missing values at λ = 0.3 and 0.7 at every n (@tbl-cellcov, @fig-cov).
Chaining the final imputations brought coverage to nominal; the extra-residual control was nominal at n = 100 but
over-covered at n ≥ 300. Both frequentist methods were close to nominal throughout, including the
fixed-parameter one, which confirms that per-value coverage checks calibration rather than properness. At λ = 1
all methods were near nominal and chained BACE slightly conservative. pigauto's posterior MI covered single values
like proper frequentist MI at n = 300 and 1000 (`r f3(cn1("pig_post", 300))` and `r f3(cn1("pig_post", 1000))`
against `r f3(cn1("freqA", 300))` and `r f3(cn1("freqA", 1000))`, pooled over λ and ρ), with the same z-RMSE
(supplement). At n = 100 it covered less at λ ≤ 0.7 (paired difference `r f3(pp("cell", 100, 0.3))` and
`r f3(pp("cell", 100, 0.7))`, SE `r f3(max(pp("cell", 100, c(0.3, 0.7), "se")))`) and more at λ = 1
(`r f3(pp("cell", 100, 1))`).

```{r}
#| label: tbl-cellcov
#| tbl-cap: "Coverage of the Rubin interval for single missing values of c1 and c2 (Monte Carlo SE), by n, λ and ρ. Nominal 0.95."
knitr::kable(wide(cells, "coverage", "coverage_se", rows = c("n", "lambda", "rho"), arms = setdiff(ord, "complete")), align = "l", row.names = FALSE)
```

```{r}
#| label: fig-cov
#| fig-cap: "Coverage of single missing values by n and λ (circles ρ = 0, triangles ρ = 0.5); whiskers ±2 Monte Carlo SE. Dashed line: nominal 0.95."
#| fig-width: 9
#| fig-height: 4.2
s <- cells[cells$arm %in% c("freqA", "freqB", "bace", "bace_chain", "bace_resid", "pig_post"), ]
figcov(s, s$coverage, s$coverage_se, c(0.85, 1))
```

The proper frequentist method was the more accurate at n = 100 and at λ = 1. At n ≥ 300 and λ ≤ 0.7 chained BACE
was more accurate (@tbl-zpair), probably because it also conditions on the count and discrete traits, which the
frequentist model leaves out.

```{r}
#| label: tbl-zpair
#| tbl-cap: "Paired per-dataset difference in z-RMSE of single missing values (chained BACE minus proper frequentist; SE over datasets), pooled over ρ. Negative favours chained BACE."
knitr::kable(data.frame(n = zp$n, `λ` = zp$lambda, difference = paste0(f3(zp$diff), " (", f3(zp$se), ")"), datasets = zp$k, check.names = FALSE)[order(zp$n, zp$lambda), ], align = "l", row.names = FALSE)
```

### Downstream estimates

Per-value coverage alone cannot show improperness, so the downstream estimates are the direct evidence
(@tbl-down, @fig-down). For the phylogenetic correlation at λ ≤ 0.7, BACE as shipped covered ρ in
`r rng(down$coverage[down$arm == "bace" & down$estimand == "cor" & down$lambda < 1 & down$n >= 300], f2)` of
datasets at n ≥ 300, and the fixed-parameter frequentist method in
`r rng(down$coverage[down$arm == "freqB" & down$estimand == "cor" & down$lambda < 1 & down$n >= 300], f2)`.
Chained BACE and proper frequentist MI covered 0.94 or more at λ ≤ 0.7; chained BACE was conservative at n = 1000
(correlation `r rng(down$coverage[down$arm == "bace_chain" & down$estimand == "cor" & down$lambda < 1 & down$n == 1000], f2)`
at λ ≤ 0.7) with wider intervals. At λ = 1, ρ = 0.5 and n = 1000 both covered the correlation only
`r f3(dv("bace_chain", 1000, 1, 0.5, "cor"))` (their slope intervals covered `r f3(dv("bace_chain", 1000, 1, 0.5, "slope"))`
and `r f3(dv("freqA", 1000, 1, 0.5, "slope"))`, @tbl-down), chained BACE through attenuation and the frequentist
method through the shared-λ bias described below. The shortfall of BACE as shipped was small per value at λ = 1 but
still visible downstream: at n = 300 it covered the correlation `r f3(dv("bace", 300, 1, 0, "cor"))` and
`r f3(dv("bace", 300, 1, 0.5, "cor"))` at ρ = 0 and 0.5, against `r f3(dv("bace_chain", 300, 1, 0, "cor"))` and
`r f3(dv("bace_chain", 300, 1, 0.5, "cor"))` chained.

With correlated traits (ρ = 0.5), chained BACE pulled the slope toward zero at n ≤ 300: bias
`r rng(dv("bace_chain", 100, c(0.3, 0.7, 1), 0.5, "slope", "bias"))` at n = 100 and
`r rng(dv("bace_chain", 300, c(0.3, 0.7, 1), 0.5, "slope", "bias"))` at n = 300 (@tbl-down), against at most
`r f3(max(abs(dv("freqA", c(100, 300, 1000), c(0.3, 0.7), 0.5, "slope", "bias"))))` in absolute value for proper
frequentist MI at λ ≤ 0.7. Chaining shifted the slope down relative to BACE as shipped in every ρ = 0.5 scenario, by
`r rng(-cshift)`. The frequentist methods were biased too at λ = 1:
their shared λ is pulled below 1 by prp's extra noise, which biased the slope by `r rng(down$bias[down$arm == "freqA" & down$lambda == 1 & down$rho == 0.5 & down$estimand == "slope"])` at ρ = 0.5 at every n while the complete data were unbiased. The extra-residual control under-covered the correlation at ρ = 0.5 and λ ≤ 0.7 at n ≥ 300 (`r rng(down$coverage[down$arm == "bace_resid" & down$estimand == "cor" & down$rho == 0.5 & down$lambda < 1 & down$n == 1000], f2)` at n = 1000), because its added noise attenuates the association. The complete-data slope interval itself covered
`r rng(dv("complete", c(100, 300), c(0.3, 0.7, 1), 0.5, "slope"))` at ρ = 0.5 and n ≤ 300, plausibly because λ is
estimated (not tested).

```{r}
#| label: tbl-down
#| tbl-cap: "Rubin-pooled PGLS slope of c2 on c1 at ρ = 0.5, by n and λ: Each cell: coverage of ρ (Monte Carlo SE); bias [datasets scored]. The correlation and ρ = 0 are in the supplement."
s <- down[down$estimand == "slope" & down$rho == 0.5 & down$arm %in% ord, ]
s$cellv <- paste0(f3(s$coverage), " (", f3(cse(s$coverage, s$datasets)), "); ", f3(s$bias), " [", s$datasets, "]")
w <- stats::reshape(s[, c("n", "lambda", "arm", "cellv")], idvar = c("n", "lambda"), timevar = "arm", direction = "wide")
names(w) <- sub("^cellv\\.", "", names(w)); keep <- intersect(ord, names(w)); w <- w[order(w$n, w$lambda), c("n", "lambda", keep)]
names(w) <- c("n", "λ", lab[keep])
knitr::kable(w, align = "l", row.names = FALSE)
```

```{r}
#| label: fig-down
#| fig-cap: "Coverage of ρ by the Rubin-pooled phylogenetic correlation, by n and λ (circles ρ = 0, triangles ρ = 0.5); whiskers ±2 Monte Carlo SE."
#| fig-width: 9
#| fig-height: 4.2
s <- down[down$estimand == "cor" & down$arm %in% c("complete", "freqA", "freqB", "bace", "bace_chain", "pig_post"), ]
figcov(s, s$coverage, cse(s$coverage, s$datasets), c(0.7, 1))
```

### pigauto's posterior MI with complete phylogenetic signal {#sec-pig}

At λ ≤ 0.7 pigauto's posterior MI covered the slope and the correlation about as often as proper frequentist MI at
every n (paired differences in coverage `r rng(pp(c("slope", "cor"), c(100, 300, 1000), c(0.3, 0.7)))`, pooled over
ρ; the largest, `r f3(pp("cor", 100, 0.7))` with SE `r f3(pp("cor", 100, 0.7, "se"))`, for the correlation at
n = 100). At λ = 1 it did not (@tbl-down, @fig-down). At ρ = 0.5 its slope was too small at every n: bias
`r rng(dv("pig_post", c(100, 300, 1000), 1, 0.5, "slope", "bias"))` at n = 100 to 1000, against
`r rng(dv("freqA", c(100, 300, 1000), 1, 0.5, "slope", "bias"))` for proper frequentist MI and
`r rng(dv("complete", c(100, 300, 1000), 1, 0.5, "slope", "bias"))` for the complete data. The bias did not shrink
as the intervals narrowed, so slope coverage fell from `r f3(dv("pig_post", 100, 1, 0.5, "slope"))` at n = 100 to
`r f3(dv("pig_post", 300, 1, 0.5, "slope"))` at n = 300 and `r f3(dv("pig_post", 1000, 1, 0.5, "slope"))` at
n = 1000 (correlation `r f3(dv("pig_post", 1000, 1, 0.5, "cor"))`; proper frequentist MI
`r f3(dv("freqA", 1000, 1, 0.5, "slope"))` and `r f3(dv("freqA", 1000, 1, 0.5, "cor"))`). At ρ = 0 and n = 1000 it
covered the slope and the correlation `r rng(c(dv("pig_post", 1000, 1, 0, "slope"), dv("pig_post", 1000, 1, 0, "cor")))`. Pooled over ρ, the paired difference from proper frequentist
MI at n = 1000 and λ = 1 was `r f3(pp("slope", 1000, 1))` (SE `r f3(pp("slope", 1000, 1, "se"))`) for the slope and
`r f3(pp("cor", 1000, 1))` (SE `r f3(pp("cor", 1000, 1, "se"))`) for the correlation.

The bias comes from pigauto's prior on the residual covariance, not from its imputation of single values. We refitted
eight campaign datasets (n = 1000, ρ = 0.5; six at λ = 1, two at λ = 0.7) and kept their completed datasets
(`script/rubin_pig_cond_diag.R`). At λ = 1 the imputed values matched the exact conditional distribution under the
true model (shrinkage slope 0.999, SD ratio 1.02), and the posterior λ of c1 and c2 was 1.000. Twenty draws from that
exact distribution, pooled in the same way, gave a mean slope error against the complete data of −0.004 over the six
λ = 1 datasets, against
−0.034 for pigauto and −0.013 for proper frequentist MI. The prior Σ~E~ ~ IW(K + 1, 0.01 diag(*s*^2^)) keeps the
residual covariance above zero, its true value at λ = 1, and pulls its correlation toward zero. The imputed entries
therefore carry a small residual noise (variance about 10^−4^ per trait) that is nearly uncorrelated between c1 and
c2. At λ = 1 and n = 1000 the slope rests on differences between sister species of about that size, so the noise
attenuates it: adding pigauto's posterior Σ~E~ as noise to the exact draws moved their slope error from −0.004 to
−0.028. At ρ = 0 there is no association to attenuate, and at λ < 1 Σ~E~ is real and estimated, which matches where
the bias appeared. Smaller prior scales for Σ~E~ removed most of the bias at n = 1000 but worsened convergence at
n = 100 in a screen of ten datasets per scenario, so none was adopted; pigauto's documentation states the caveat
(pigauto pull request #201). pigauto's conformal draws, which are not meant for downstream inference, were tried
only in a 36-fit pre-run and are not a method of this study: they biased the slope by −0.08 to −0.15 and covered it in
half of the 12 datasets at each of n = 300 and 1000.

## Discrete traits

### Accuracy and calibration of imputed classes

At λ = 0.3 and 0.7 chained BACE imputed all three discrete traits more accurately than castor (@tbl-dacc,
@fig-dacc): by `r rng(lead, f2)` on the same datasets, in every scenario
(`r rng(leadz, function(x) formatC(x, format = "f", digits = 1))` SEs). At λ = 0.3 castor was less accurate than
simply guessing the most frequent observed class for every species (by
`r rng(-(dcl$accuracy[dcl$arm == "freqA" & dcl$lambda == 0.3] - dcl$accuracy[dcl$arm == "mode_floor" & dcl$lambda == 0.3]))`).
castor's flexible fits there have very large rates, so its draws follow the model's stationary distribution and
carry almost no phylogenetic information; with equal rates that distribution is uniform, which cost up to
`r f2(max(dpr$diff[dpr$contrast == "freqA - freqA_er" & dpr$measure == "accuracy" & dpr$lambda < 1]))` accuracy.
BACE's lead was `r rng(pa$diff[pa$lambda < 1 & pa$rho == 0], f2)` at ρ = 0, where the other traits are
uncorrelated with the imputed one but informative jointly with d1, and
`r rng(pa$diff[pa$lambda < 1 & pa$rho == 0.5], f2)` at ρ = 0.5. At λ = 1 castor was as accurate or slightly more
so (paired differences `r rng(pa$diff[pa$lambda == 1])`).

```{r}
#| label: tbl-dacc
#| tbl-cap: "Accuracy of the most frequent imputed class (Monte Carlo SE), pooled over ρ. Higher is better. The last column always guesses the most frequent observed class."
knitr::kable(wide_d(dcl, function(g) cell_se(g$accuracy, g$accuracy_se), c("trait", "n", "lambda"),
                    c("freqA", "freqB", "freqA_er", "bace", "bace_chain", "mode_floor")), align = "l", row.names = FALSE)
```

```{r}
#| label: fig-dacc
#| fig-cap: "Accuracy of the most frequent imputed class at n = 300 by trait and λ, pooled over ρ; whiskers ±2 Monte Carlo SE. Grey bars: always guessing the most frequent observed class."
#| fig-width: 9
#| fig-height: 4.2
s <- dcl[dcl$n == 300 & dcl$arm %in% c("freqA", "freqB", "bace", "bace_chain"), ]
fl <- dcl[dcl$n == 300 & dcl$arm == "mode_floor", ]
s$trait <- factor(TRN[s$trait], levels = TRN); fl$trait <- factor(TRN[fl$trait], levels = TRN)
s$method <- factor(lab[s$arm], levels = lab)
ggplot(s, aes(factor(lambda), accuracy, colour = method)) +
  geom_crossbar(data = fl, aes(factor(lambda), accuracy, ymin = accuracy, ymax = accuracy), inherit.aes = FALSE,
                width = 0.8, colour = "#b9c2bd", linewidth = 1.1) +
  geom_pointrange(aes(ymin = accuracy - 2 * accuracy_se, ymax = accuracy + 2 * accuracy_se),
                  position = position_dodge(width = 0.6), size = 0.35, linewidth = 0.6) +
  facet_wrap(~trait) + scale_colour_manual(values = pal, name = NULL) +
  labs(x = "phylogenetic signal λ", y = "accuracy of most frequent imputed class") + theme_rubin
```

The 95% class sets and Brier scores (supplement) repeated the continuous pattern. At λ = 0.3 and n = 100 the sets
of BACE as shipped covered
`r rng(dcl$set_coverage[dcl$arm == "bace" & dcl$lambda == 0.3 & dcl$n == 100], f2)` and chained BACE
`r rng(dcl$set_coverage[dcl$arm == "bace_chain" & dcl$lambda == 0.3 & dcl$n == 100], f2)`, below the 0.95 a
calibrated 20-draw imputer reaches, and chaining lowered the Brier score in every λ ≤ 0.7 scenario (by
`r rng(-dpr$diff[dpr$contrast == "bace_chain - bace" & dpr$measure == "brier" & dpr$lambda < 1])`).

### A downstream association with a binary trait

Imputing each trait without the other biased their association (@tbl-ddown, @fig-ddown). At ρ = 0.5 and λ ≤ 0.7
the proper frequentist datasets pulled the slope of c1 on bin toward zero by
`r rng(-ddv("freqA", c(0.3, 0.7), 0.5, c(100, 300), "bias_cond"), f2)`. The bias did not shrink with n while the
standard error did, so coverage fell from `r rng(ddv("freqA", c(0.3, 0.7), 0.5, 100, "coverage_cond"), f2)` at
n = 100 to `r rng(ddv("freqA", c(0.3, 0.7), 0.5, 300, "coverage_cond"), f2)` at n = 300. At ρ = 0, where the true
slope is zero, the same datasets biased the slope upward by
`r rng(ddv("freqA", c(0.3, 0.7), 0, c(100, 300), "bias_cond"), f2)`. In plain terms: c1 and bin are each linked to
the always-measured trait d1, so imputing c1 from d1 alone, and bin from the tree alone, creates a spurious positive
link between them in the imputed species. (Formally, their correlation given d1 is −0.14, and imputing each without
the other sets it to zero.) Their intervals nevertheless covered
`r rng(ddv("freqA", c(0.3, 0.7), 0, c(100, 300), "coverage_cond"), f3)`, because when the imputation model and the
analysis model disagree about which variables are related, Rubin's variance no longer measures the true sampling
variance; here it overstated it (this mismatch is called uncongeniality; Meng 1994). BACE imputes bin given c1: at λ ≤ 0.7 its bias was
`r rng(ddv("bace", c(0.3, 0.7), c(0, 0.5), c(100, 300), "bias_cond"))` (chained
`r rng(ddv("bace_chain", c(0.3, 0.7), c(0, 0.5), c(100, 300), "bias_cond"))`) and it covered
`r rng(ddv("bace", c(0.3, 0.7), c(0, 0.5), c(100, 300), "coverage_cond"))` as shipped and
`r rng(ddv("bace_chain", c(0.3, 0.7), c(0, 0.5), c(100, 300), "coverage_cond"))` chained, against
`r rng(ddv("complete", c(0.3, 0.7), c(0, 0.5), c(100, 300), "coverage_cond"))` for the complete data. At λ = 1
and ρ = 0.5 the ranking reversed: the slope from BACE as shipped was `r f3(-ddv("bace", 1, 0.5, 100, "bias_cond"))` too small at
n = 100 (SE `r f3(ddv("bace", 1, 0.5, 100, "bias_cond_se"))`), more than castor's
(`r f3(-ddv("freqA", 1, 0.5, 100, "bias_cond"))` proper, `r f3(-ddv("freqB", 1, 0.5, 100, "bias_cond"))` fixed
parameters; chained BACE `r f3(-ddv("bace_chain", 1, 0.5, 100, "bias_cond"))`), and all methods over-covered. A few BACE fits there had very large between-imputation variance (mean
interval width up to about 100), so interval widths are not compared.

```{r}
#| label: tbl-ddown
#| tbl-cap: "PGLS slope of c1 on bin: coverage of the per-dataset target by the Rubin-pooled 95% interval, and bias of the pooled estimate (Monte Carlo SE). Compare coverage with the complete-data column."
da2 <- c("complete", "freqA", "freqB", "bace", "bace_chain")
cvt <- wide_d(ddn, function(g) cell_se(g$coverage_cond, g$coverage_cond_se), c("n", "lambda", "rho"), da2)
bst <- wide_d(ddn, function(g) cell_se(g$bias_cond, g$bias_cond_se), c("n", "lambda", "rho"), c("freqA", "freqB", "bace", "bace_chain"))
names(cvt)[-(1:3)] <- paste("coverage,", names(cvt)[-(1:3)]); names(bst)[4:7] <- paste("bias,", names(bst)[4:7])
knitr::kable(cbind(cvt, bst[, 4:7]), align = "l", row.names = FALSE)
```

```{r}
#| label: fig-ddown
#| fig-cap: "Coverage of the per-dataset target for the slope of c1 on bin at ρ = 0.5, by n and λ; whiskers ±2 Monte Carlo SE. Dashed line: 0.95."
#| fig-width: 9
#| fig-height: 4.2
s <- ddn[ddn$rho == 0.5 & ddn$arm %in% c("complete", "freqA", "freqB", "bace", "bace_chain"), ]
s$n <- s$n; figcov(s, s$coverage_cond, cse(s$coverage_cond, s$n_fits), c(0.2, 1), shape = FALSE)
```

## Failures

BACE stopped with MCMCglmm's "Mixed model equations singular" on `r length(cfail)` of 3,000 continuous-part
datasets and `r length(bfail)` of 2,400 discrete-part datasets, all at λ = 1; chained BACE failed on
`r chain_only_c` and `r chain_only_d` further datasets. At n ≤ 300, `r length(intersect(bfail, cfail300))` of the
`r length(cfail300)` continuous-part failures recurred in the discrete run; `r length(setdiff(union(bfail, cfail300), intersect(bfail, cfail300)))`
datasets failed in only one run, one of them on the same cluster both times. These counts are final failures. A
first pass of BACE, before the input cleaning described in @sec-methods, left
`r sum(failures$files[failures$set == "bace_asshipped_failed"])` failure records
(`r paste(paste0(failures$files[failures$set == "bace_asshipped_failed"], " at n = ", failures$n[failures$set == "bace_asshipped_failed"], ", λ = ", failures$lambda[failures$set == "bace_asshipped_failed"]), collapse = "; ")`);
those datasets were refitted after cleaning, and the final failures are the ones that still stopped. Failed fits are recorded and not
scored, so BACE's rows at λ = 1 rest on fewer datasets than the others (the tables give the counts); the datasets
BACE failed on were, if anything, slightly easier for the frequentist method. The frequentist methods fitted every
dataset apart from `r sum(failures$freq_errors)` Rphylopars failures in the continuous part. pigauto's sampler
stopped with a Cholesky error (a matrix not positive definite) on `r sum(pf$pig_errors)` of `r big(sum(pf$files))`
datasets (`r paste(paste0(pf$pig_errors[pf$pig_errors > 0], " at n = ", pf$n[pf$pig_errors > 0], ", λ = ", pf$lambda[pf$pig_errors > 0]), collapse = "; ")`),
which are not scored. A further `r sum(pf$pig_unconverged)` pigauto fits, all at n = `r paste(unique(pf$n[pf$pig_unconverged > 0]), collapse = ", ")`
and λ ≤ 0.7, still failed its convergence rule after the last extension; they are scored and flagged, and at n = 100
they covered at least as often as the converged fits (0.96 to 0.98 against 0.91 to 0.96), so they do not explain
its lower coverage there. BACE returned no class
for about 1.6% of its three-class draws (non-finite class probabilities); scores use the remaining draws. BACE's
own convergence check tests its deterministic settling phase and does not measure mixing, so it was not used to
choose settings.

# Discussion

BACE's imputations as shipped are not proper. Every final imputation starts from the same dataset, whose missing
entries were filled with posterior means rather than draws, so each imputation conditions on one fixed version of
the other traits' missing values. The draws for single values are too narrow, and the between-imputation variance
misses the uncertainty in those fills; the downstream correlation, whose coverage fell to
`r f2(dv("bace", 1000, 0.7, 0, "cor"))` at n = 1000, λ = 0.7 and ρ = 0, shows it directly. Chaining the final
imputations is a small change that removes the shared start. Chained BACE then covered at or above nominal
(conservatively at n = 1000) except for the correlation at λ = 1, ρ = 0.5 and n = 1000, and, at n ≥ 300 and λ ≤ 0.7, was more accurate per value than proper frequentist MI,
though it attenuated the slope with correlated traits, most at small n and at λ = 1. Its successive imputations are consecutive sweeps
of one chain, so independent chains, or several sweeps between retained imputations, may remove the remaining
dependence.

The two parts of the study favour different methods for reasons that belong beside the results. The continuous
traits follow the model Rphylopars fits, apart from prp's extra noise, which biased the frequentist slope at
λ = 1. The discrete traits are thresholded liabilities, the model family BACE fits, whereas castor treats each trait
alone with an Mk process; cat3 was generated as an ordered trait and modelled as nominal by both methods. BACE's
advantage on discrete traits therefore mixes two sources, conditioning on the other traits and matching the data
model, and this design cannot separate them.

pigauto's posterior MI shows that a proper scheme can still be biased by its prior. Its imputations of single
values matched the exact conditional distribution, yet a residual-covariance prior that cannot reach zero
attenuated the slope where the true residual variance is zero (λ = 1) and the tree is large enough (n = 1000) for
the slope to rest on small differences between close relatives.

The downstream result generalises beyond these tools. An imputation model that omits a variable of the analysis is
uncongenial with it (Meng 1994): it pulls the association toward the value the imputer assumes (here zero), Rubin's
variance cannot correct that bias, and more data make the interval more confidently wrong. The recommendations in
the box after the Introduction follow from these results.

## Limitations

- One mechanism: Brownian motion with Pagel's λ on random coalescent trees, MCAR 30%. Missing-at-random or
  clade-structured missingness, Ornstein-Uhlenbeck evolution and real trees were not studied, and no empirical
  example is included.
- The frequentist continuous model shares one λ across c1, c2, logit(prp) and d1. prp's extra noise pulls that λ
  below 1 at λ = 1, which biased the frequentist slope and correlation by −0.018 to −0.026 at ρ = 0.5 at every n; at λ = 1 and ρ = 0.5 the proper method's slope coverage was `r f3(dv("freqA", 100, 1, 0.5, "slope"))`,
  `r f3(dv("freqA", 300, 1, 0.5, "slope"))` and `r f3(dv("freqA", 1000, 1, 0.5, "slope"))` at n = 100, 300 and 1000,
  and its correlation coverage fell to `r f3(dv("freqA", 1000, 1, 0.5, "cor"))` at n = 1000. The model also omits the count and discrete traits.
- Chained BACE and the extra-residual control were designed after a pre-run and tested on the same seeds.
- pigauto's posterior MI was added after the main campaign and imputes the continuous block only. Its λ = 1
  diagnosis rests on eight refitted datasets, and the prior trials on eight datasets at n = 1000 and twenty at
  n = 100.
- BACE at n = 1000 used 100 datasets per scenario (Monte Carlo SEs about 1.4 times the others'). The discrete part
  covers n = 100 and 300 only, so its trends in n rest on two sample sizes.
- The discrete traits follow BACE's model family, and nominal traits generated from several liabilities were not
  simulated. An intercept-only BACE model and Mk-generated traits would separate model matching from conditioning.
- At λ = 1 about half of BACE's fits needed input cleaning, and failed fits are not a random subset.
- At λ = 1 and ρ = 0.5 the target of the slope of c1 on bin varies between datasets by more than its standard
  error, so coverage there is read against the per-dataset target.

# Data and code availability

All results, tables and code are on branch `arc/rubin-freq-bace` of the pigauto repository (the R package in which
this study was run). This document is built from the tables in `data/`:

| file | contents |
|---|---|
| `data/fits.csv`, `data/discrete/fits.csv` | one row per fit: scenario, seed, cluster, node, time, wall times, errors, software versions |
| `data/fit_cells.csv.gz`, `data/fit_estimands.csv.gz` | continuous part: per fit × method × trait (per-value scores) and × estimand (pooled estimates) |
| `data/discrete/fit_disc_cells.csv.gz`, `fit_disc_estimands.csv.gz` | discrete part: per fit × method × trait and × estimand, with both targets |
| `data/agg_pig/`, `data/discrete/agg_disc_*.csv` | aggregated tables used here; `data/agg_pig/` adds the pigauto rows to `data/agg/`, whose rows it reproduces exactly |
| `data/discrete/truth_slope_c1_bin.csv` | Monte Carlo population target of the slope of c1 on bin |
| `data/discrete/repro.csv` | reproduction of the stored continuous fits by the discrete-part re-run |
| `data/bace_diagnostics.csv`, `data/discrete/bace_diagnostics.csv` | BACE convergence verdicts, effective sample sizes, input cleaning |
| `data/failure_records.csv`, `data/discrete/castor_diag.csv.gz` | first-pass BACE failures; castor fit diagnostics |

Code: `script/rubin_cell.R` generates and imputes one dataset with all methods; `script/campaign_gnn_off_lib.R`
(`make_cell()`), `script/rubin_freq.R`, `script/rubin_bace.R`, `script/rubin_discrete.R` and `script/rubin_lib.R`
hold data generation, methods and pooling, with `script/rubin_pigauto.R` for pigauto's posterior MI; `script/rubin_campaign_aggregate.R`, `script/rubin_discrete_aggregate.R`
and `script/rubin_study/build_data.R` build the tables; tests are in `script/tests-rubin/`. The raw per-fit result
files, including every imputation of the discrete part (1.4 GB), are archived separately.

# Supplement {.unnumbered}

```{r}
#| label: tbl-s-down
#| tbl-cap: "Supplementary: Rubin-pooled downstream estimates by n, λ and ρ: coverage of ρ (Monte Carlo SE), bias and datasets scored."
s <- down[down$arm %in% ord, ]; s <- s[order(s$estimand, s$n, s$lambda, s$rho, match(s$arm, ord)), ]
knitr::kable(data.frame(estimand = ifelse(s$estimand == "slope", "slope", "correlation"), n = s$n, `λ` = s$lambda, `ρ` = s$rho, method = lab[s$arm],
                        coverage = paste0(f3(s$coverage), " (", f3(cse(s$coverage, s$datasets)), ")"), bias = paste0(f3(s$bias), " (", f3(s$bias_se), ")"),
                        width = f3(s$ci_width), FMI = f3(s$fmi_median), datasets = s$datasets, check.names = FALSE), align = "l", row.names = FALSE)
```

```{r}
#| label: tbl-s-acc
#| tbl-cap: "Supplementary: z-RMSE of single missing values of c1 and c2 (SD units; Monte Carlo SE), pooled over ρ."
knitr::kable(wide(cells_l, "zrmse", "zrmse_se", arms = setdiff(ord, "complete")), align = "l", row.names = FALSE)
```

```{r}
#| label: tbl-s-dset
#| tbl-cap: "Supplementary: Brier score and coverage of the 95% class set of the 20 draws (Monte Carlo SE), pooled over ρ."
b <- wide_d(dcl, function(g) cell_se(g$brier, g$brier_se), c("trait", "n", "lambda"), c("freqA", "bace", "bace_chain"))
cv <- wide_d(dcl, function(g) cell_se(g$set_coverage, g$set_coverage_se), c("trait", "n", "lambda"), c("freqA", "bace", "bace_chain"))
names(b)[4:6] <- paste("Brier,", names(b)[4:6]); names(cv)[4:6] <- paste("set coverage,", names(cv)[4:6])
knitr::kable(cbind(b, cv[, 4:6]), align = "l", row.names = FALSE)
```

```{r}
#| label: tbl-s-time
#| tbl-cap: "Supplementary: wall time per fit (one dataset, all methods of the set) on one core."
w <- rbind(data.frame(part = "continuous", aggregate(I(wall_s_total / 3600) ~ set + n, fits, function(x) c(median = median(x), p90 = unname(quantile(x, 0.9))))),
           data.frame(part = "discrete", aggregate(I(wall_s_total / 3600) ~ set + n, dfits, function(x) c(median = median(x), p90 = unname(quantile(x, 0.9))))))
w <- data.frame(part = w$part, set = w$set, n = w$n, round(w[[4]], 2)); names(w)[4:5] <- c("median hours", "90th percentile hours")
knitr::kable(w, align = "l", row.names = FALSE)
```

```{r}
#| label: tbl-s-fail
#| tbl-cap: "Supplementary: continuous part by n and λ. files: result files; bace_errors: BACE fits that stopped; freq_errors: Rphylopars fits that failed; bace_cleaned: BACE fits whose inputs needed empty levels dropped or a trait left out; freqA refit columns: bootstrap refits redrawn. Rows labelled bace_asshipped_failed are first-pass BACE runs, before input cleaning, kept as records. Rows labelled pig are pigauto's posterior MI; pig_errors: fits that stopped; pig_unconverged: fits scored but flagged as unconverged."
knitr::kable(failures, align = "l", row.names = FALSE)
```

# References {.unnumbered}

- Barnard, J. & Rubin, D. B. (1999). Small-sample degrees of freedom with multiple imputation. *Biometrika* 86, 948–955.
- Brier, G. W. (1950). Verification of forecasts expressed in terms of probability. *Monthly Weather Review* 78, 1–3.
- Gneiting, T. & Raftery, A. E. (2007). Strictly proper scoring rules, prediction, and estimation. *Journal of the American Statistical Association* 102, 359–378.
- Goolsby, E. W., Bruggeman, J. & Ané, C. (2017). Rphylopars: fast multivariate phylogenetic comparative methods for missing data and within-species variation. *Methods in Ecology and Evolution* 8, 22–27.
- Hadfield, J. D. (2010). MCMC methods for multi-response generalized linear mixed models: the MCMCglmm R package. *Journal of Statistical Software* 33(2), 1–22.
- Louca, S. & Doebeli, M. (2018). Efficient comparative phylogenetics on large trees. *Bioinformatics* 34, 1053–1055.
- Meng, X.-L. (1994). Multiple-imputation inferences with uncongenial sources of input. *Statistical Science* 9, 538–558.
- Morris, T. P., White, I. R. & Crowther, M. J. (2019). Using simulation studies to evaluate statistical methods. *Statistics in Medicine* 38, 2074–2102.
- Nakagawa, S. & Freckleton, R. P. (2008). Missing inaction: the dangers of ignoring missing data. *Trends in Ecology & Evolution* 23, 592–596.
- Pagel, M. (1999). Inferring the historical patterns of biological evolution. *Nature* 401, 877–884.
- Paradis, E. & Schliep, K. (2019). ape 5.0: an environment for modern phylogenetics and evolutionary analyses in R. *Bioinformatics* 35, 526–528.
- Rubin, D. B. (1987). *Multiple Imputation for Nonresponse in Surveys*. Wiley, New York.
- Williams et al. (2024). Transparent reporting items for simulation studies evaluating statistical methods. *Methods in Ecology and Evolution* 15, 1926–1939. (Author list to be checked before submission.)
