---
title:  "phyloimpute: Simulated & Real‑Data Walk‑Through"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{phyloimpute Tutorial (Simulated + AVONET)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7)
#library(pigauto)
library(ape)
library(dplyr)
library(ggplot2)
library(phytools)
```

# 1  Introduction

`phyloimpute` fills missing trait values by borrowing information from  
*phylogenetic relatives* **and** *environmental covariates* through three deep‑learning approaches:

| Method | Strength | Uncertainty |
|--------|----------|-------------|
| **VGAE** (Variational Graph AE) | uses the tree explicitly | latent sampling |
| **MC Dropout AE** | simple, fast | Monte‑Carlo dropout variance |
| **Ensemble AE** | robust aggregation | between‑model variance |

We illustrate both a **simulated** dataset (to verify recovery of known values) and a **real** subset of the **AVONET** bird trait database.

---

# 2  Simulation Example

## 2.1  Simulate phylogeny + traits

```{r simulate-data}
set.seed(42)
tree_sim  <- rtree(60)
trait_mat <- data.frame(
  Mass      = fastBM(tree_sim, sig2 = 1),
  BeakLen   = fastBM(tree_sim, sig2 = 0.5),
  Diet      = factor(sample(c("Insect","Fruit","Seed"), 60, TRUE)),
  Habitat   = factor(sample(c("Forest","Grass","Desert"), 60, TRUE))
)
env_sim <- data.frame(
  Temp = rnorm(60, 15, 7),
  NDVI = runif(60, 0.3, 0.8)
)
```

## 2.2  Mask 20 % at random

```{r mask}
mask_idx <- sample(length(unlist(trait_mat)), 0.2*length(unlist(trait_mat)))
trait_na <- trait_mat
trait_na[mask_idx] <- NA
mean(is.na(trait_na))
```

## 2.3  Impute with the three methods

```{r impute-sim, eval = FALSE}
imp_vgae  <- impute_vgae(trait_na, tree_sim, env_sim, epochs = 80)
imp_drop  <- impute_mc_dropout(trait_na, env_sim, epochs = 80, mc = 30)
imp_ens   <- impute_ensemble(trait_na, env_sim, epochs = 80, n_models = 5)
```

## 2.4  Accuracy for one trait

```{r accuracy, eval = FALSE}
truth <- trait_mat$Mass[is.na(trait_na$Mass)]
predV <- imp_vgae $Mass[is.na(trait_na$Mass)]
predD <- imp_drop $Mass[is.na(trait_na$Mass)]
predE <- imp_ens  $Mass[is.na(trait_na$Mass)]
data.frame(Method = c("VGAE","Drop","Ens"),
           RMSE   = sqrt(c(mean((predV-truth)^2),
                           mean((predD-truth)^2),
                           mean((predE-truth)^2))))
```

## 2.5  Visualise uncertainty (MC Dropout)

```{r uncert-plot, eval = FALSE}
unc <- attr(imp_drop, "sd")[, "Mass"]
ggplot(data.frame(Truth = trait_mat$Mass, Pred = imp_drop$Mass, SD = unc),
       aes(Truth, Pred, colour = SD)) +
  geom_point(size = 2) + geom_abline(lty = 2) +
  scale_colour_viridis_c() + theme_minimal()
```

---

# 3  Real‑Data Example (AVONET + WorldClim)

```{r load-real}
data(avonet300); data(env300); data(tree300)
dim(avonet300)
```

## 3.1  Create additional missingness

```{r miss-real}
set.seed(21)
na_rows <- sample(nrow(avonet300), 30)
avonet_miss <- avonet300
avonet_miss$WingLen[na_rows] <- NA
```

## 3.2  Run VGAE

```{r vgae-real, eval = FALSE}
imp_vgae_real <- impute_vgae(avonet_miss, tree300, env300, epochs = 100)
```

## 3.3  Compare observed vs imputed distribution

```{r density, eval = FALSE}
ggplot() +
  geom_density(aes(avonet300$WingLen), fill = "grey80") +
  geom_density(aes(imp_vgae_real$WingLen), colour = "blue") +
  labs(x = "Wing length (mm)", y = "Density",
       title = "Observed vs imputed distribution (VGAE)")
```

## 3.4  Timing summary

```{r timing, eval = FALSE}
bench::mark(
  VGAE    = impute_vgae(avonet_miss, tree300, env300, epochs = 60),
  Dropout = impute_mc_dropout(avonet_miss, env300, epochs = 60),
  Ensemble= impute_ensemble(avonet_miss, env300, epochs = 60, n_models = 3),
  iterations = 1
)[,1:5]
```

---

# 4  Saving a trained model

```{r save-model, eval = FALSE}
fit <- impute_vgae(avonet_miss, tree300, env300, epochs = 120)
saveRDS(fit, "vgae_birds.rds")
fit2 <- readRDS("vgae_birds.rds")
```

---

# 5  Session info

```{r session}
sessionInfo()
```
