---
title: "phyloimpute: Extensive Tutorial"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{phyloimpute Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
#library(phyloimpute)
library(ape)
library(ggplot2)
library(dplyr)
library(phytools)
```

# 1 Introduction

`phyloimpute` fills missing ecological trait data by combining *phylogenetic
relationships* and *environmental covariates* in deep autoencoder models.
This vignette shows **end‑to‑end usage** on a simulated dataset:

1. Simulate a phylogeny, traits, and environment  
2. Introduce missing values under different patterns  
3. Impute with the three methods (`vgae`, `mc_dropout`, `bootstrap`)  
4. Inspect completed data and uncertainty  
5. Evaluate performance (because truth is known)  
6. Visualise latent space & uncertainty heatmaps  

> **Note:** Internal autoencoder functions are *placeholders* until the full
torch implementations are added (see package README).  
Still, the workflow below illustrates **how you will interact with the final
package** — once the models are in place, the same code will produce real
imputations and uncertainties.

# 2 Simulating Data

```{r simulate}
set.seed(2025)

# 2.1 Phylogeny with 40 species
n_sp  <- 40
tree  <- rtree(n_sp)
tree$tip.label <- paste0("sp", seq_len(n_sp))

# 2.2 Continuous traits with phylogenetic signal (Brownian motion)
trait_cont1 <- fastBM(tree, sig2 = 1)
trait_cont2 <- fastBM(tree, sig2 = 0.5)

# 2.3 Ordinal trait derived from latent BM + noise
latent_ord  <- fastBM(tree)
breaks      <- quantile(latent_ord, probs = c(0.33, 0.66))
trait_ord   <- cut(latent_ord,
                   breaks = c(-Inf, breaks, Inf),
                   labels = c("small", "medium", "large"),
                   ordered_result = TRUE)

# 2.4 Nominal trait correlated with clade (simulate using tip numbers)
trait_nom   <- factor(ifelse(as.numeric(sub("sp","",tree$tip.label)) <= 20,
                             "A", "B"))

# 2.5 Binary presence/absence depending on trait_cont1
prob_bin    <- plogis(trait_cont1)
trait_bin   <- rbinom(n_sp, 1, prob_bin)

# 2.6 Environmental covariate (numeric temperature & categorical biome)
env_temp    <- rnorm(n_sp, 15, 5)
env_biome   <- factor(sample(c("forest","grassland"), n_sp, TRUE))

# 2.7 Assemble data frames
traits_full <- data.frame(
  species = tree$tip.label,
  mass    = trait_cont1,
  length  = trait_cont2,
  sizecat = trait_ord,
  clade   = trait_nom,
  presence= factor(trait_bin, labels = c("absent","present"))
)

env_full <- data.frame(
  temp  = env_temp,
  biome = env_biome
)

head(traits_full)
```

## 2.8 Introduce Missingness

We remove 25 % of entries completely at random **plus** all `mass`
values for one sub‑clade (species 31–40) to mimic a real clade gap.

```{r missingness}
traits_miss <- traits_full
# random 25% NAs
set.seed(1)
miss_mat <- matrix(runif(prod(dim(traits_miss[,-1]))) < 0.25,
                   nrow=n_sp)
for(j in seq_len(ncol(traits_miss)-1)) {
  traits_miss[-1][[j]][ miss_mat[,j] ] <- NA
}
# clade gap for mass
traits_miss$mass[31:40] <- NA

summary(is.na(traits_miss))
```

# 3 Imputation Workflows

We demonstrate **all three methods**.  
*(With placeholders each method currently falls back to simple column means or modes;
once torch models are implemented, they will run full training loops).*

```{r impute}
species_id <- traits_miss$species

res_vgae  <- impute_phylo(traits_miss[,-1], tree,
                          env_data = env_full,
                          species_id = species_id,
                          method = "vgae",
                          latent_dim = 8,
                          epochs = 100)

res_drop  <- impute_phylo(traits_miss[,-1], tree,
                          env_data = env_full,
                          species_id = species_id,
                          method = "mc_dropout")

res_boot  <- impute_phylo(traits_miss[,-1], tree,
                          env_data = env_full,
                          species_id = species_id,
                          method = "bootstrap")
```

# 4 Completed Data

```{r completed}
completed <- res_vgae$completed_data
head(completed)
```

# 5 Quick Accuracy Check (Continuous traits)

Because we know the truth, let’s compute RMSE for `mass` after imputation.

```{r rmse}
rmse <- sqrt(mean((completed$mass - traits_full$mass)^2))
cat("RMSE (mass) placeholder:", rmse, "\n")
```

With full models you would expect a **substantial reduction** vs. mean
imputation, especially for the clade with all missing data.

# 6 Visualising Imputed vs. True (mass)

```{r scatter}
ggplot(data.frame(true = traits_full$mass,
                  imp  = completed$mass,
                  group = ifelse(is.na(traits_miss$mass), "missing","observed")),
       aes(true, imp, colour = group)) +
  geom_point(size = 2) +
  geom_abline(linetype = "dashed") +
  theme_minimal() +
  labs(title = "Imputed vs. true (continuous trait 'mass')")
```

# 7 Uncertainty Heatmap

```{r heatmap, eval=FALSE}
# enable once uncertainty implemented
# plot_uncertainty(res_vgae)
```

# 8 Latent Space (VGAE)

When the variational graph autoencoder is fully implemented, you will be
able to extract and plot the latent embedding (e.g. first two dims) as a
`ggplot2` scatter coloured by traits or environment to inspect clustering.

# 9 Runtime Benchmarks

```{r benchmark, eval=FALSE}
# timings <- bench::mark(
#   vgae = impute_phylo(...),
#   dropout = impute_phylo(...),
#   bootstrap = impute_phylo(..., n_models=5)
# )
# timings
```

# 10 Next Steps

The placeholder functions demonstrate the *interface*.  
To deploy production‑grade imputation:

1. Implement full torch training loops in `models_vgae.R`,
   `models_dropout.R`, and `models_ensemble.R`.
2. Add trait‑type decoding logic (continuous, ordinal, nominal, binary).
3. Re‑build and run this vignette — all analysis and visualisations will
   use the **true model outputs**.

For details, see package README and function documentation.

```{r session}
sessionInfo()
```
