---
title: "Four ways to impute a phylogenetic trait matrix"
subtitle: "A simulation study comparing the frequentist stack, BACE, and pigauto with the GNN off and on"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Four ways to impute a phylogenetic trait matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE,
                      fig.width = 7.2, fig.height = 4.2, dpi = 120)
# Pre-rendered from committed aggregate csv. Nothing is computed when the site builds: the campaign
# behind this page is thousands of core-hours and cannot run in a pkgdown job.
res_dir <- system.file("simstudy", package = "pigauto")
if (!nzchar(res_dir)) res_dir <- file.path("..", "..", "script", "campaign_sim_results")
have <- file.exists(file.path(res_dir, "summary.csv"))
if (have) {
  S <- read.csv(file.path(res_dir, "summary.csv"), stringsAsFactors = FALSE)
  P <- tryCatch(read.csv(file.path(res_dir, "paired.csv"), stringsAsFactors = FALSE),
                error = function(e) NULL)
  FL <- tryCatch(read.csv(file.path(res_dir, "failures.csv"), stringsAsFactors = FALSE),
                 error = function(e) NULL)
}
ARM <- c(freq = "Frequentist, BM default", freq_lambda = "Frequentist, lambda estimated",
         bace = "BACE",
         gnn_off = "pigauto, GNN off", gnn_off_rphylopars = "pigauto, GNN off, Rphylopars solver",
         gnn_on = "pigauto, GNN on", floor = "Mean / mode floor")
KEEP <- c("freq", "freq_lambda", "bace", "gnn_off", "gnn_off_rphylopars", "gnn_on", "floor")
fmt <- function(m, s) ifelse(is.finite(m), sprintf("%.3f (%.3f)", m, s), "n/a")
```

```{r not-yet, eval = !exists("have") || !have, results = "asis"}
cat("> **This page is a placeholder.** The campaign behind it has not finished, so no numbers are",
    "shown yet. The design below is fixed and was registered before any result was inspected.\n")
```

> **Scope of the numbers on this page.** Everything below is the *core slice*: Brownian motion,
> MCAR at 30%, no covariates, lambda in {0.3, 0.7, 1.0}, rho in {0, 0.5}, n in {100, 300, 1000},
> 200 replicates per cell and 100 for BACE. The factorial extension over the evolutionary model,
> the missingness mechanism and the missing fraction is still computing and will be added here
> rather than replacing anything. The design was fixed before any result was inspected.

## What this compares, and why

Four ways of filling gaps in a species-by-trait matrix, scored against a known truth.

1. **The frequentist stack.** `Rphylopars` for a joint fit over the
   continuous-family traits, `castor`'s Mk model for each discrete trait, and a phylogenetic Poisson
   GLM for counts. The standard toolkit, with no information passing between trait types. It is
   reported twice, because the choice of evolutionary model turns out to matter more than the
   choice of method: once at Rphylopars' documented default `model = "BM"`, and once at
   `model = "lambda"`, which estimates the phylogenetic signal from the data.
2. **BACE.** Joint Bayesian imputation through `MCMCglmm`, which models all traits together and
   returns posterior predictive draws.
3. **pigauto with the GNN off.** The phylogenetic baseline and pigauto's safety machinery, with no
   neural network. Reported with both of its joint solvers.
4. **pigauto with the GNN on.** The shipped default.

A mean or mode floor sits underneath, so that a method doing worse than ignoring the phylogeny
entirely is visible as such rather than merely looking mediocre.

The question is not only which is most accurate. An imputation you plan to analyse afterwards has to
be honest about its own uncertainty, so accuracy, probability calibration and interval coverage are
weighed together, with the cost of each fit recorded beside them.

## How the data were made

Species evolve on an ultrametric tree from `ape::rcoal`, rescaled to unit height. Latent traits come
from a multivariate normal whose phylogenetic part is Pagel's lambda applied to the tree correlation,

$$V_\lambda = \lambda V + (1-\lambda) I,$$

with unit diagonal, so the marginal variance of every latent is exactly one at every lambda. A
second matrix gives every pair of traits a cross-trait correlation rho. Eight traits are built from
those latents: two continuous, a Poisson count, a proportion, a binary trait, a four-level ordinal, a
three-level categorical, and one always-observed continuous trait that drives the
missing-at-random mechanism.

Discrete traits are cut at **fixed population thresholds** rather than sample quantiles, so class
balance varies between replicates. That is realistic, and it exercises the rare-class regime where
imputation methods actually fail; sample quantiles would force perfect balance in every replicate.

Cells then go missing in one of four ways: completely at random at 10% and 30%; at random at 30%,
driven by the always-observed trait with an odds ratio of 3 per standard deviation; and
phylogenetically biased at 30%, where whole clades are unsampled. Every arm in a replicate sees the
same mask, so all comparisons are paired.

```{r design-table, eval = exists("have") && have}
knitr::kable(data.frame(
  Factor = c("Species", "Phylogenetic signal", "Cross-trait correlation", "Missingness",
             "Evolutionary model", "Replicates"),
  Levels = c("100, 300, 1000", "lambda 0.3, 0.7, 1.0", "rho 0, 0.5",
             "MCAR 10%, MCAR 30%, MAR 30%, clade-biased 30%", "Brownian motion, Ornstein-Uhlenbeck",
             "200 per cell; BACE on the first 100, nested")),
  caption = "The design. The core slice is Brownian motion with MCAR 30%.")
```

## Accuracy

```{r accuracy, eval = exists("have") && have}
core <- S[S$evo == "BM" & S$miss == "mcar" & S$metric == "zRMSE" &
            S$frac == 0.3 & S$trait %in% c("c1", "c2", "cnt", "prp"), ]
if (nrow(core)) {
  agg <- aggregate(cbind(mean, mcse) ~ n + lambda + arm, data = core,
                   FUN = function(z) mean(z, na.rm = TRUE))
  agg$cell <- sprintf("n = %s, lambda = %s", agg$n, agg$lambda)
  w <- reshape(agg[, c("cell", "arm", "mean")], idvar = "cell", timevar = "arm", direction = "wide")
  names(w) <- sub("^mean\\.", "", names(w))
  keep <- intersect(KEEP, names(w))
  knitr::kable(w[, c("cell", keep)], digits = 3, row.names = FALSE,
               col.names = c("Cell", ARM[keep]),
               caption = "z-RMSE on masked cells, averaged over the continuous-family traits. Lower is better. Monte Carlo standard errors are in the accompanying csv and on every figure.")
}
```

Read this table with the floor column in view. An arm above the floor is doing worse than predicting
the sample mean, which is the honest description of a phylogenetic method applied where there is
little phylogenetic signal to exploit.

```{r accuracy-discrete, eval = exists("have") && have}
cored <- S[S$evo == "BM" & S$miss == "mcar" & S$frac == 0.3 & S$metric == "accuracy" &
             S$trait %in% c("bin", "ord", "cat3"), ]
if (nrow(cored)) {
  agg <- aggregate(mean ~ n + lambda + arm, data = cored, FUN = function(z) mean(z, na.rm = TRUE))
  agg$cell <- sprintf("n = %s, lambda = %s", agg$n, agg$lambda)
  w <- reshape(agg[, c("cell", "arm", "mean")], idvar = "cell", timevar = "arm", direction = "wide")
  names(w) <- sub("^mean\\.", "", names(w))
  keep <- intersect(KEEP, names(w))
  knitr::kable(w[, c("cell", keep)], digits = 3, row.names = FALSE,
               col.names = c("Cell", ARM[keep]),
               caption = "Accuracy on masked cells, averaged over the binary, ordinal and three-level categorical traits. Higher is better.")
}
```

### What the two tables say

The ranking is not stable across the design, and that is the main result rather than a complication.

Phylogenetic signal decides who wins on continuous traits. At lambda = 0.3 every arm that assumes
maximum signal sits at or above the floor, and the two that estimate the signal, BACE and the
frequentist stack at `model = "lambda"`, are the only ones clearly below it. At lambda = 1 the order
inverts: pigauto with the GNN off reaches 0.379 at n = 1000 against BACE's 0.659, and BACE is last.

The frequentist stack's evolutionary model matters more here than the choice of method. Moving
Rphylopars from its `model = "BM"` default to `model = "lambda"` improves z-RMSE by about 0.23 at
lambda = 0.3 and by about 0.18 at lambda = 0.7, which is larger than the gap between most pairs of
methods in the same cell. At the default the stack sits above the floor at lambda = 0.3; with the
signal estimated it never does.

Discrete traits run the other way. BACE is ahead by 11 to 17 accuracy points wherever signal is
low or moderate, and every other arm is at or below the mode floor at lambda = 0.3. At lambda = 1
that reverses and the Mk and label-propagation paths reach 0.96 against BACE's 0.84.

The two pigauto solvers are not interchangeable. The in-house solver and the Rphylopars solver
agree in most cells, but the Rphylopars solver is unstable at high signal and small n: at lambda = 1
and n = 100 it reaches a z-RMSE of 1.654 against the in-house solver's 0.487, and its interval score
diverges in two cells. That is a reason to keep the in-house solver as the default.

## Calibration and coverage

The three arms do not build the same kind of interval, and the comparison says so rather than
quietly treating them as equivalent:

- `Rphylopars` gives a **model-based** interval for an unobserved tip, with the Brownian parameters
  treated as known.
- BACE gives a **Bayesian posterior predictive** interval, taken as percentiles over its final
  imputed datasets.
- pigauto gives a **split-conformal** interval, whose distribution-free guarantee holds over cells
  exchangeable with its calibration split. Under clade-biased missingness the masked cells are not
  exchangeable with the observed ones, so that guarantee does not transfer; the results below show
  what happens when it does not.

Coverage is never reported without the standardised width beside it, and the interval score of
Gneiting and Raftery is given as the single proper score that trades the two against each other.

### What the coverage table says

Only one of the three interval constructions reaches its nominal level. pigauto's split-conformal
intervals cover 0.95 to 0.96 at n = 300 and n = 1000, across every lambda. BACE's posterior
predictive intervals reach 0.80 to 0.89 and are furthest from nominal where the signal is weakest.
The model-based Rphylopars intervals fall between the two, at 0.80 to 0.91.

The pattern has a simple reading. The conformal interval is calibrated on held-out residuals, so it
absorbs whatever error the model actually makes; the other two are derived from a fitted model and
inherit its optimism. The exception proves it: at n = 100 pigauto drops to 0.85 to 0.88, because a
100-tip tree leaves too few validation residuals for the conformal quantile to be stable. Coverage
is not a free property of being Bayesian or of being frequentist, and on this evidence it is bought
by calibrating against held-out error.

```{r coverage, eval = exists("have") && have}
cv <- S[S$evo == "BM" & S$miss == "mcar" & S$frac == 0.3 &
          S$metric %in% c("coverage", "width", "interval_score") &
          S$arm %in% KEEP & S$trait %in% c("c1", "c2", "cnt", "prp"), ]
if (nrow(cv)) {
  agg <- aggregate(mean ~ arm + metric + n, data = cv, FUN = function(z) mean(z, na.rm = TRUE))
  w <- reshape(agg, idvar = c("arm", "n"), timevar = "metric", direction = "wide")
  names(w) <- sub("^mean\\.", "", names(w))
  w$arm <- ARM[w$arm]
  knitr::kable(w, digits = 3, row.names = FALSE,
               caption = "Nominal 95% coverage, mean standardised width and interval score, by sample size.")
}
```

## Cost

```{r cost, eval = exists("have") && have}
knitr::kable(data.frame(
  Arm = c("Frequentist stack", "pigauto, GNN off", "pigauto, GNN off, Rphylopars solver",
          "pigauto, GNN on", "BACE"),
  `n = 100` = c("0.4 s", "0.7 s", "27 s", "134 s", "950 s"),
  `n = 1000` = c("2 s", "38 s", "260 s", "319 s", "2 h 46 m"),
  check.names = FALSE),
  caption = "Median wall time for one fit on one core, measured during this campaign.")
```

Cost is part of the comparison, not an aside. The Bayesian arm is roughly three orders of magnitude
more expensive than the frequentist stack at n = 100, and that ratio widens with sample size. Whether
it buys anything is what the accuracy and coverage tables are for.

## Failures

```{r failures, eval = exists("have") && have && exists("FL") && !is.null(FL) && nrow(FL)}
knitr::kable(aggregate(n_failed ~ arm, data = FL[FL$arm %in% KEEP, ], FUN = sum),
             caption = "Fits that failed outright, by arm. A failed fit is scored at the floor and counted here, never dropped.")
```

Methods differ in where they break, and that belongs in the comparison. BACE's mixed model equations
become singular under clade-biased missingness, which is exactly the regime a real phylogenetic
dataset with unsampled clades presents.

## Reproducing this

Runners, the design table and the aggregated results are in the pigauto repository under `script/`.
Each replicate stores its latent matrix, realised parameters, seeds, random number generator kind and
the package versions used, and the generator is fixed so that a replicate reproduces across machines.

The full methods description, written for the accompanying paper, is in
`docs/dev-log/arc/2026-09-20-simulation-methods-bace.md`.

## What this study does not cover

Multi-observation designs, multi-proportion and zero-inflated traits, and missing-not-at-random
mechanisms are outside its scope. The Ornstein-Uhlenbeck arm is a sensitivity check on model
misspecification, not a full treatment. The comparison uses each method as its documentation
recommends, and does not tune any of them beyond those defaults.
