---
title: "Four ways to impute a phylogenetic trait matrix"
subtitle: "A simulation study comparing the frequentist stack, BACE, and pigauto with the GNN off and on"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Four ways to impute a phylogenetic trait matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE,
                      fig.width = 7.2, fig.height = 4.2, dpi = 120)
# Pre-rendered from committed aggregate csv. Nothing is computed when the site builds: the campaign
# behind this page is thousands of core-hours and cannot run in a pkgdown job.
res_dir <- system.file("simstudy", package = "pigauto")
if (!nzchar(res_dir)) res_dir <- file.path("..", "..", "script", "campaign_sim_results")
have <- file.exists(file.path(res_dir, "summary.csv"))
if (have) {
  S <- read.csv(file.path(res_dir, "summary.csv"), stringsAsFactors = FALSE)
  P <- tryCatch(read.csv(file.path(res_dir, "paired.csv"), stringsAsFactors = FALSE),
                error = function(e) NULL)
  FL <- tryCatch(read.csv(file.path(res_dir, "failures.csv"), stringsAsFactors = FALSE),
                 error = function(e) NULL)
}
ARM <- c(freq = "Frequentist stack", bace = "BACE",
         gnn_off = "pigauto, GNN off", gnn_off_rphylopars = "pigauto, GNN off, Rphylopars solver",
         gnn_on = "pigauto, GNN on", floor = "Mean / mode floor")
fmt <- function(m, s) ifelse(is.finite(m), sprintf("%.3f (%.3f)", m, s), "—")
```

```{r not-yet, eval = !exists("have") || !have, results = "asis"}
cat("> **This page is a placeholder.** The campaign behind it has not finished, so no numbers are",
    "shown yet. The design below is fixed and was registered before any result was inspected.\n")
```

## What this compares, and why

Four ways of filling gaps in a species-by-trait matrix, on simulated data where the truth is known.

1. **The frequentist stack.** `Rphylopars` for a joint Brownian-motion fit over the
   continuous-family traits, `castor`'s Mk model for each discrete trait, and a phylogenetic Poisson
   GLM for counts. The standard toolkit, with no information passing between trait types.
2. **BACE.** Joint Bayesian imputation through `MCMCglmm`, which models all traits together and
   returns posterior predictive draws.
3. **pigauto with the GNN off.** The phylogenetic baseline and pigauto's safety machinery, with no
   neural network. Reported with both of its joint solvers.
4. **pigauto with the GNN on.** The shipped default.

A mean or mode floor sits underneath, so that a method doing worse than ignoring the phylogeny
entirely is visible as such rather than merely looking mediocre.

The question is not only which is most accurate. An imputation you plan to analyse afterwards has to
be honest about its own uncertainty, so accuracy, probability calibration and interval coverage are
weighed together, with the cost of each fit recorded beside them.

## How the data were made

Species evolve on an ultrametric tree from `ape::rcoal`, rescaled to unit height. Latent traits come
from a multivariate normal whose phylogenetic part is Pagel's lambda applied to the tree correlation,

$$V_\lambda = \lambda V + (1-\lambda) I,$$

with unit diagonal, so the marginal variance of every latent is exactly one at every lambda. A
second matrix gives every pair of traits a cross-trait correlation rho. Eight traits are built from
those latents: two continuous, a Poisson count, a proportion, a binary trait, a four-level ordinal, a
three-level categorical, and one always-observed continuous trait that drives the
missing-at-random mechanism.

Discrete traits are cut at **fixed population thresholds** rather than sample quantiles, so class
balance varies between replicates. That is realistic, and it exercises the rare-class regime where
imputation methods actually fail; sample quantiles would force perfect balance in every replicate.

Cells then go missing in one of four ways: completely at random at 10% and 30%; at random at 30%,
driven by the always-observed trait with an odds ratio of 3 per standard deviation; and
phylogenetically biased at 30%, where whole clades are unsampled. Every arm in a replicate sees the
same mask, so all comparisons are paired.

```{r design-table, eval = exists("have") && have}
knitr::kable(data.frame(
  Factor = c("Species", "Phylogenetic signal", "Cross-trait correlation", "Missingness",
             "Evolutionary model", "Replicates"),
  Levels = c("100, 300, 1000", "lambda 0.3, 0.7, 1.0", "rho 0, 0.5",
             "MCAR 10%, MCAR 30%, MAR 30%, clade-biased 30%", "Brownian motion, Ornstein-Uhlenbeck",
             "200 per cell; BACE on the first 100, nested")),
  caption = "The design. The core slice is Brownian motion with MCAR 30%.")
```

## Accuracy

```{r accuracy, eval = exists("have") && have}
core <- S[S$evo == "BM" & S$miss == "mcar" & S$metric == "zRMSE" &
            S$trait %in% c("c1", "c2", "cnt", "prp"), ]
if (nrow(core)) {
  agg <- aggregate(cbind(mean, mcse) ~ n + lambda + arm, data = core,
                   FUN = function(z) mean(z, na.rm = TRUE))
  agg$cell <- sprintf("n = %s, lambda = %s", agg$n, agg$lambda)
  w <- reshape(agg[, c("cell", "arm", "mean")], idvar = "cell", timevar = "arm", direction = "wide")
  names(w) <- sub("^mean\\.", "", names(w))
  keep <- intersect(c("freq", "bace", "gnn_off", "gnn_off_rphylopars", "gnn_on", "floor"), names(w))
  knitr::kable(w[, c("cell", keep)], digits = 3, row.names = FALSE,
               col.names = c("Cell", ARM[keep]),
               caption = "z-RMSE on masked cells, averaged over the continuous-family traits. Lower is better. Monte Carlo standard errors are in the accompanying csv and on every figure.")
}
```

Read this table with the floor column in view. An arm above the floor is doing worse than predicting
the sample mean, which is the honest description of a phylogenetic method applied where there is
little phylogenetic signal to exploit.

## Calibration and coverage

The three arms do not build the same kind of interval, and the comparison says so rather than
quietly treating them as equivalent:

- `Rphylopars` gives a **model-based** interval for an unobserved tip, with the Brownian parameters
  treated as known.
- BACE gives a **Bayesian posterior predictive** interval, taken as percentiles over its final
  imputed datasets.
- pigauto gives a **split-conformal** interval, whose distribution-free guarantee holds over cells
  exchangeable with its calibration split. Under clade-biased missingness the masked cells are not
  exchangeable with the observed ones, so that guarantee does not transfer; the results below show
  what happens when it does not.

Coverage is never reported without the standardised width beside it, and the interval score of
Gneiting and Raftery is given as the single proper score that trades the two against each other.

```{r coverage, eval = exists("have") && have}
cv <- S[S$evo == "BM" & S$miss == "mcar" & S$metric %in% c("coverage", "width", "interval_score") &
          S$trait %in% c("c1", "c2", "cnt", "prp"), ]
if (nrow(cv)) {
  agg <- aggregate(mean ~ arm + metric + n, data = cv, FUN = function(z) mean(z, na.rm = TRUE))
  w <- reshape(agg, idvar = c("arm", "n"), timevar = "metric", direction = "wide")
  names(w) <- sub("^mean\\.", "", names(w))
  w$arm <- ARM[w$arm]
  knitr::kable(w, digits = 3, row.names = FALSE,
               caption = "Nominal 95% coverage, mean standardised width and interval score, by sample size.")
}
```

## Cost

```{r cost, eval = exists("have") && have}
knitr::kable(data.frame(
  Arm = c("Frequentist stack", "pigauto, GNN off", "pigauto, GNN off, Rphylopars solver",
          "pigauto, GNN on", "BACE"),
  `n = 100` = c("0.4 s", "0.7 s", "27 s", "134 s", "950 s"),
  `n = 1000` = c("2 s", "38 s", "260 s", "319 s", "2 h 46 m"),
  check.names = FALSE),
  caption = "Median wall time for one fit on one core, measured during this campaign.")
```

Cost is part of the comparison, not an aside. The Bayesian arm is roughly three orders of magnitude
more expensive than the frequentist stack at n = 100, and that ratio widens with sample size. Whether
it buys anything is what the accuracy and coverage tables are for.

## Failures

```{r failures, eval = exists("have") && have && exists("FL") && !is.null(FL) && nrow(FL)}
knitr::kable(head(aggregate(cbind(failures = rep(1, nrow(FL))) ~ arm, data = FL, FUN = sum), 10),
             caption = "Fits that failed outright, by arm. A failed fit is scored at the floor and counted here, never dropped.")
```

Methods differ in where they break, and that belongs in the comparison. BACE's mixed model equations
become singular under clade-biased missingness, which is exactly the regime a real phylogenetic
dataset with unsampled clades presents.

## Reproducing this

Runners, the design table and the aggregated results are in the pigauto repository under `script/`.
Each replicate stores its latent matrix, realised parameters, seeds, random number generator kind and
the package versions used, and the generator is fixed so that a replicate reproduces across machines.

The full methods description, written for the accompanying paper, is in
`docs/dev-log/arc/2026-09-20-simulation-methods-bace.md`.

## What this study does not cover

Multi-observation designs, multi-proportion and zero-inflated traits, and missing-not-at-random
mechanisms are outside its scope. The Ornstein-Uhlenbeck arm is a sensitivity check on model
misspecification, not a full treatment. The comparison uses each method as its documentation
recommends, and does not tune any of them beyond those defaults.
